---
title: "17june"
output: html_document
date: "2025-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




To construct animal models for your pearl oyster traits using ASReml in R (as in your lizard example), you'll need:

Your phenotypic dataset (dat)

A pedigree file with columns for ID, Sire, and Dam

The inverse of the additive relationship matrix (A⁻¹)

Here's a generalised structure you can adapt for each significant GLM trait, beginning with an example for the binary Gold trait using Pearl_Harvest_Data_DVH_mm.


```{r}
# Clear workspace and load libraries
rm(list = ls())
library(asreml)
library(nadiv)

# Read your phenotypic and pedigree data
dat <- read.csv("your_data.csv")  # Replace with your file
ped <- dat[, c("Animal_ID", "Sire_ID", "Dam_ID")]  # Match your column names

# Convert columns to factors
dat$Animal_ID <- as.factor(dat$Animal_ID)
ped <- data.frame(
  id = as.factor(ped[, 1]),
  sire = as.factor(ped[, 2]),
  dam = as.factor(ped[, 3])
)

# Build inverse relationship matrix
ainv <- ainverse(ped)

# Example: Animal model for Gold ~ Pearl_Harvest_Data_DVH_mm
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Animal_ID, ainv),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)
mod.gold <- update.asreml(mod.gold)

# View results
summary(mod.gold)$varcomp
vpredict(mod.gold, h2 ~ V1 / (V1 + V2))  # Heritability estimate

```


 Repeat for Other Traits
Just change the fixed effect and the trait. Example for Silver:

```{r}
mod.silver <- asreml(
  fixed = Silver ~ First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(Animal_ID, ainv),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)
mod.silver <- update.asreml(mod.silver)

```

Same pattern for White, Cream, Champ—always use the relevant fixed predictors from your GLM results.










#######################

1. Load libraries and Data

```{r}
library(readxl)
library(nadiv)
library(asreml)
library(readr)

# Load pedigree data
ped_host <- read_csv("~/pearl_project/animalmodel/host.csv")
View(ped_host)

ped_donor <- read_csv("~/pearl_project/animalmodel/saibo.csv")
View(ped_donor)

# Load phenotypic data
PearlOperationalData <- read_csv("~/pearl_project/PearlOperationalData.csv")
View(PearlOperationalData)

dat <- PearlOperationalData

```


2. Format Pedigree and Phenotypic Data

```{r}
# Format pedigree for host
ped_host <- data.frame(
  id = as.factor(ped_host$Shell_Metrics_GenotypeID_DART),
  sire = as.factor(ped_host$Sire),
  dam = as.factor(ped_host$Dam)
)

# Format pedigree for donor
# Remove the "18-03_Saibo_" prefix from IDs
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)

ped_donor <- data.frame(
  id = as.factor(ped_donor$Matched_Donor_Data_Saibo_No),
  sire = as.factor(ped_donor$Sire),
  dam = as.factor(ped_donor$Dam)
)

# Format phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)  # Host IDs

# Match the donor ID formatting in phenotype data to the pedigree
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)

```



3. Generate A inverse Matrix 
```{r}
ainv_host <- ainverse(ped_host)


ainv_donor <- ainverse(ped_donor)

```


4.    Prepare binary trait (Gold)

Made sure Gold is coded as 0/1

```{r}
dat$Gold <- as.integer(dat$Pearl_Grading_Data_Colour == "Gold")

```


5. Fit animal model: Gold ~ DVH

```{r}
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)

mod.gold <- update.asreml(mod.gold)
summary(mod.gold)$varcomp
vpredict(mod.gold, h2 ~ V1 / (V1 + V2))  # Heritability estimate

```




