---
title: "17june"
output: html_document
date: "2025-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```




To construct animal models for your pearl oyster traits using ASReml in R (as in your lizard example), you'll need:

Your phenotypic dataset (dat)

A pedigree file with columns for ID, Sire, and Dam

The inverse of the additive relationship matrix (A⁻¹)

Here's a generalised structure you can adapt for each significant GLM trait, beginning with an example for the binary Gold trait using Pearl_Harvest_Data_DVH_mm.


```{r}
# Clear workspace and load libraries
rm(list = ls())
library(asreml)
library(nadiv)

# Read your phenotypic and pedigree data
dat <- read.csv("your_data.csv")  # Replace with your file
ped <- dat[, c("Animal_ID", "Sire_ID", "Dam_ID")]  # Match your column names

# Convert columns to factors
dat$Animal_ID <- as.factor(dat$Animal_ID)
ped <- data.frame(
  id = as.factor(ped[, 1]),
  sire = as.factor(ped[, 2]),
  dam = as.factor(ped[, 3])
)

# Build inverse relationship matrix
ainv <- ainverse(ped)

# Example: Animal model for Gold ~ Pearl_Harvest_Data_DVH_mm
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Animal_ID, ainv),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)
mod.gold <- update.asreml(mod.gold)

# View results
summary(mod.gold)$varcomp
vpredict(mod.gold, h2 ~ V1 / (V1 + V2))  # Heritability estimate

```


 Repeat for Other Traits
Just change the fixed effect and the trait. Example for Silver:

```{r}
mod.silver <- asreml(
  fixed = Silver ~ First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(Animal_ID, ainv),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)
mod.silver <- update.asreml(mod.silver)

```

Same pattern for White, Cream, Champ—always use the relevant fixed predictors from your GLM results.










#######################

1. Load libraries and Data

```{r}
library(readxl)
library(nadiv)
library(asreml)
library(readr)

# Load pedigree data
ped_host <- read_csv("~/pearl_project/animalmodel/host.csv")
View(ped_host)

ped_donor <- read_csv("~/pearl_project/animalmodel/saibo.csv")
View(ped_donor)

# Load phenotypic data
PearlOperationalData <- read_csv("~/pearl_project/PearlOperationalData.csv")
View(PearlOperationalData)

dat <- PearlOperationalData

```


2. Format Pedigree and Phenotypic Data

```{r}
# Format pedigree for host
ped_host <- data.frame(
  id = as.factor(ped_host$Shell_Metrics_GenotypeID_DART),
  sire = as.factor(ped_host$Sire),
  dam = as.factor(ped_host$Dam)
)

# Format pedigree for donor
# Remove the "18-03_Saibo_" prefix from IDs
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)

ped_donor <- data.frame(
  id = as.factor(ped_donor$Matched_Donor_Data_Saibo_No),
  sire = as.factor(ped_donor$Sire),
  dam = as.factor(ped_donor$Dam)
)

# Format phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)  # Host IDs

# Match the donor ID formatting in phenotype data to the pedigree
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)

```



3. Generate A inverse Matrix 
```{r}
ainv_host <- ainverse(ped_host)


ainv_donor <- ainverse(ped_donor)

```


4.    Prepare binary trait (Gold)

Made sure Gold is coded as 0/1

```{r}
dat$Gold <- as.integer(dat$Pearl_Grading_Data_Colour == "Gold")

```


5. Fit animal model: Gold ~ DVH

```{r}
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)

mod.gold <- update.asreml(mod.gold)
summary(mod.gold)$varcomp
vpredict(mod.gold, h2 ~ V1 / (V1 + V2))  # Heritability estimate

```



########################################




```{r}
# 1. Load Libraries
library(readxl)
library(nadiv)
library(asreml)
library(readr)

# 2. Load Data ------------------------------------------------------

# Load host pedigree
ped_host <- read_csv("~/pearl_project/animalmodel/host.csv")
# Load donor pedigree
ped_donor <- read_csv("~/pearl_project/animalmodel/saibo.csv")
# Load phenotype data
PearlOperationalData <- read_csv("~/pearl_project/PearlOperationalData.csv")

# Assign to working variable
dat <- PearlOperationalData

# 3. Format Pedigrees ------------------------------------------------------

# Format host pedigree
ped_host <- data.frame(
  id = as.factor(ped_host$Shell_Metrics_GenotypeID_DART),
  sire = as.factor(ped_host$Sire),
  dam = as.factor(ped_host$Dam)
)

# Format donor pedigree (remove prefix)
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)
ped_donor <- data.frame(
  id = as.factor(ped_donor$Matched_Donor_Data_Saibo_No),
  sire = as.factor(ped_donor$Sire),
  dam = as.factor(ped_donor$Dam)
)

# 4. Format Phenotypic Data -----------------------------------------------

# Format host and donor IDs in phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)

# Binary trait (Gold)
dat$Gold <- as.integer(dat$Pearl_Grading_Data_Colour == "Gold")

# 5. Clean and Subset Data -----------------------------------------------

# Remove NAs in response or predictor
dat <- dat[!is.na(dat$Gold) & !is.na(dat$Pearl_Harvest_Data_DVH_mm), ]

# Keep only individuals in phenotype that are in the pedigree
dat <- dat[dat$Shell_Metrics_GenotypeID_DART %in% ped_host$id, ]

# Double-check factor format
dat$Shell_Metrics_GenotypeID_DART <- factor(dat$Shell_Metrics_GenotypeID_DART)

# 6. Generate Inverse A Matrices -----------------------------------------

ainv_host <- ainverse(ped_host)
ainv_donor <- ainverse(ped_donor)  # Optional: for future dual random effect models

# 7. Fit Animal Model (Gold ~ DVH) ---------------------------------------

mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family = asr_binomial(),
  data = dat
)

# Update model (for convergence)
mod.gold <- update.asreml(mod.gold)

# 8. Output Results ------------------------------------------------------

# Variance components
summary(mod.gold)$varcomp

# Heritability estimate
vpredict(mod.gold, h2 ~ V1 / (V1 + V2))

```





######################################################################



```{r}
# 1. Load Libraries ------------------------------------------------------
library(readxl)
library(nadiv)
library(asreml)
library(readr)

# 2. Load Data -----------------------------------------------------------

# Load host pedigree
ped_host <- read_csv("~/pearl_project/animalmodel/host.csv")

# Load donor pedigree
ped_donor <- read_csv("~/pearl_project/animalmodel/saibo.csv")

# Load phenotype data
PearlOperationalData <- read_csv("~/pearl_project/PearlOperationalData.csv")

# Assign to working variable
dat <- PearlOperationalData


# 3. Format Pedigrees ---------------------------------------------------

# Format host pedigree
ped_host <- data.frame(
  id   = as.factor(ped_host$Shell_Metrics_GenotypeID_DART),
  sire = as.factor(ped_host$Sire),
  dam  = as.factor(ped_host$Dam)
)

# Format donor pedigree (remove prefix)
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)
ped_donor <- data.frame(
  id   = as.factor(ped_donor$Matched_Donor_Data_Saibo_No),
  sire = as.factor(ped_donor$Sire),
  dam  = as.factor(ped_donor$Dam)
)


# 4. Format Phenotypic Data ---------------------------------------------

# Clean and format IDs
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)

# Check available levels in Pearl_Grading_Data_Colour
cat("Unique pearl colours:\n")
print(unique(dat$Pearl_Grading_Data_Colour))

# Create binary Gold trait
dat$Gold <- as.integer(dat$Pearl_Grading_Data_Colour == "Gold")

# Check for variation in Gold trait
cat("Gold counts:\n")
print(table(dat$Gold, useNA = "ifany"))



# 4. Format Phenotypic Data ---------------------------------------------

# Clean and format IDs
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)

# Clean colour data and create binary Gold trait
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Check that there are 0s and 1s
cat("Gold counts:\n")
print(table(dat$Gold, useNA = "ifany"))



# 5. Clean and Subset Data ----------------------------------------------

# Remove rows with missing trait or predictor
dat <- dat[!is.na(dat$Gold) & !is.na(dat$Pearl_Harvest_Data_DVH_mm), ]

# Retain only matched individuals in pedigree
dat <- dat[dat$Shell_Metrics_GenotypeID_DART %in% ped_host$id, ]
dat$Shell_Metrics_GenotypeID_DART <- droplevels(dat$Shell_Metrics_GenotypeID_DART)

print(table(dat$Gold, useNA = "ifany"))

# 6. Generate Inverse A Matrices ----------------------------------------

ainv_host <- ainverse(ped_host)
ainv_donor <- ainverse(ped_donor)  # optional for future use


# 7. Fit Animal Model (Gold ~ DVH) --------------------------------------

mod.gold <- asreml(
  fixed    = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random   = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family   = asr_binomial(),
  data     = dat
)

# Ensure convergence
mod.gold <- update.asreml(mod.gold)


# 8. Output Results ------------------------------------------------------

# Print variance components
cat("\nVariance components:\n")
print(summary(mod.gold)$varcomp)

# Estimate heritability (on link scale)
cat("\nHeritability estimate (link scale):\n")
print(vpredict(mod.gold, h2 ~ V1 / (V1 + V2)))

```

