---
title: "Untitled"
output: html_document
date: "2025-06-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Pearl Oyster Animal Model Code (inspired by the Lizard heritability model)

# 1. Load libraries
library(readr)
library(asreml)
library(nadiv)

# 2. Load data
ped_host <- host_Copy
ped_donor <- saibo_Copy
dat <- PearlOperationalData

# 3. Format pedigree data
ped_host <- data.frame(
  id = as.factor(ped_host$Shell_Metrics_GenotypeID_DART),
  sire = as.factor(ped_host$Sire),
  dam = as.factor(ped_host$Dam)
)
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)
ped_donor <- data.frame(
  id = as.factor(ped_donor$Matched_Donor_Data_Saibo_No),
  sire = as.factor(ped_donor$Sire),
  dam = as.factor(ped_donor$Dam)
)

# 4. Format phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# 5. Create binary traits
dat$Gold   <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White  <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream  <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ  <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 6. Inverse A matrix for host
ainv_host <- ainverse(ped_host)
ainv_donor <- ainverse(ped_donor)

# 7. Fit animal model for Gold ~ DVH
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family = asr_binomial(link = "logit"),  # Added link specification
  data = dat[!is.na(dat$Gold) & !is.na(dat$Pearl_Harvest_Data_DVH_mm), ]
)
mod.gold <- update.asreml(mod.gold)

# Calculate heritability on the liability scale
# For binary traits, we need to account for the residual variance being fixed to π²/3 in logit models
summary(mod.gold)$varcomp

# Summarize the variance components
varcomp_summary <- summary(mod.gold)$varcomp

# Extract genetic variance (VG) and residual variance (VR)
VG <- varcomp_summary["Shell_Metrics_GenotypeID_DART", "Variance"]  # Genetic variance component
VR <- varcomp_summary["units", "Variance"]  # Residual variance component

# Residual variance for binary traits (logit link) is fixed at pi^2 / 3
residual_variance_binary <- (pi^2) / 3

# Calculate heritability on the liability scale
heritability_liability_scale <- VG / (VG + residual_variance_binary)

# Print the heritability estimate
print(paste("Heritability on the liability scale:", heritability_liability_scale))




```



#doesntwork this one- Heritability calculation on liability scale
vpredict(mod.gold, h2 ~ V1/(V1 + (pi^2/3)))  # V1 is the additive genetic variance component


```{r}
# Genetic variance (VG)
VG <- 50.49  # as per the output of varcomp for the genetic variance

# Residual variance for binary traits (logit link)
residual_variance_binary <- (pi^2) / 3

# Calculate heritability on the liability scale
heritability_liability_scale <- VG / (VG + residual_variance_binary)

# Print heritability estimate
print(paste("Heritability on the liability scale:", heritability_liability_scale))

```

##The heritability on the liability scale is approximately 0.93, which indicates a very high heritability for this binary trait.



###################################################################################




```{r}
# Load libraries
library(readr)
library(asreml)
library(nadiv)

# 1. Load data
ped_host <- host_Copy
ped_donor <- saibo_Copy
dat <- PearlOperationalData

# 2. Format pedigree data
format_pedigree <- function(ped_data, id_col, sire_col, dam_col) {
  data.frame(
    id = as.factor(ped_data[[id_col]]),
    sire = as.factor(ped_data[[sire_col]]),
    dam = as.factor(ped_data[[dam_col]])
  )
}

ped_host <- format_pedigree(ped_host, "Shell_Metrics_GenotypeID_DART", "Sire", "Dam")
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)
ped_donor <- format_pedigree(ped_donor, "Matched_Donor_Data_Saibo_No", "Sire", "Dam")

# 3. Format phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# 4. Create binary traits (Gold, Silver, White, etc.)
create_binary_trait <- function(data, trait_name) {
  ifelse(startsWith(trimws(data$Pearl_Grading_Data_Colour), trait_name), 1, 0)
}

traits <- c("Gold", "Silver", "White", "Cream", "Champ")
for (trait in traits) {
  dat[[trait]] <- create_binary_trait(dat, trait)
}

# 5. Inverse A matrix for host & donor
ainv_host <- ainverse(ped_host)
ainv_donor <- ainverse(ped_donor)

# 6. Fit animal model for Gold ~ DVH
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Shell_Metrics_GenotypeID_DART, ainv_host),
  residual = ~ idv(units),
  family = asr_binomial(link = "logit"),
  data = dat[!is.na(dat$Gold) & !is.na(dat$Pearl_Harvest_Data_DVH_mm), ]
)
mod.gold <- update.asreml(mod.gold)

# 7. Extract Variance Components & Calculate Heritability
varcomp_summary <- summary(mod.gold)$varcomp
print(varcomp_summary)  # To ensure you're extracting correct components

# Extract Genetic and Residual Variance
VG <- varcomp_summary["Shell_Metrics_GenotypeID_DART", "Variance"]  # Genetic variance
VR <- varcomp_summary["units", "Variance"]  # Residual variance

# Residual Variance for Binary Traits (Logit Link)
residual_variance_binary <- (pi^2) / 3

# Calculate Heritability on the Liability Scale
heritability_liability_scale <- VG / (VG + residual_variance_binary)

# Print Heritability Estimate
print(paste("Heritability on the liability scale:", heritability_liability_scale))

```

####################################################################################


```{r}
# Load libraries
library(readr)
library(asreml)
library(nadiv)

# 1. Load data
ped_host <- host_Copy
ped_donor <- saibo_Copy
dat <- PearlOperationalData

# 2. Format pedigree data
format_pedigree <- function(ped_data, id_col, sire_col, dam_col) {
  data.frame(
    id = as.factor(ped_data[[id_col]]),
    sire = as.factor(ped_data[[sire_col]]),
    dam = as.factor(ped_data[[dam_col]])
  )
}

ped_host <- format_pedigree(ped_host, "Shell_Metrics_GenotypeID_DART", "Sire", "Dam")
ped_donor$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", ped_donor$Matched_Donor_Data_Saibo_No)
ped_donor <- format_pedigree(ped_donor, "Matched_Donor_Data_Saibo_No", "Sire", "Dam")

# 3. Format phenotypic data
dat$Shell_Metrics_GenotypeID_DART <- as.factor(dat$Shell_Metrics_GenotypeID_DART)
dat$Matched_Donor_Data_Saibo_No <- sub("18-03_Saibo_", "", dat$Matched_Donor_Data_Saibo_No)
dat$Matched_Donor_Data_Saibo_No <- as.factor(dat$Matched_Donor_Data_Saibo_No)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# 4. Create binary traits (Gold, Silver, White, etc.)
create_binary_trait <- function(data, trait_name) {
  ifelse(startsWith(trimws(data$Pearl_Grading_Data_Colour), trait_name), 1, 0)
}

traits <- c("Gold", "Silver", "White", "Cream", "Champ")
for (trait in traits) {
  dat[[trait]] <- create_binary_trait(dat, trait)
}

# 5. Inverse A matrix for host & donor
ainv_host <- ainverse(ped_host)
ainv_donor <- ainverse(ped_donor)

# 6. Fit animal model for Gold ~ DVH
mod.gold <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm,
  random = ~ vm(Matched_Donor_Data_Saibo_No, ainv_donor),
  residual = ~ idv(units),
  family = asr_binomial(link = "logit"),
  data = dat[!is.na(dat$Gold) & !is.na(dat$Pearl_Harvest_Data_DVH_mm), ]
)
mod.gold <- update.asreml(mod.gold)

# 7. Extract Variance Components & Calculate Heritability
varcomp_summary <- summary(mod.gold)$varcomp
print(varcomp_summary)  # To ensure you're extracting correct components

# Extract Genetic and Residual Variance
VG <- varcomp_summary["Matched_Donor_Data_Saibo_No", "Variance"]  # Genetic variance
VR <- varcomp_summary["units", "Variance"]  # Residual variance

# Residual Variance for Binary Traits (Logit Link)
residual_variance_binary <- (pi^2) / 3

# Calculate Heritability on the Liability Scale
heritability_liability_scale <- VG / (VG + residual_variance_binary)

# Print Heritability Estimate
print(paste("Heritability on the liability scale:", heritability_liability_scale))

```

```{r}
# Extracted Genetic Variance
VG <- 12.0666079  # Genetic variance from the output

# Residual Variance for Binary Traits (Logit Link)
residual_variance_binary <- (pi^2) / 3  # Fixed at pi^2 / 3 for binary traits

# Calculate Heritability on the Liability Scale
heritability_liability_scale <- VG / (VG + residual_variance_binary)

# Print Heritability Estimate
print(paste("Heritability on the liability scale:", heritability_liability_scale))

```
#You have successfully extracted the necessary variance components and calculated heritability on the liability scale. The heritability is about 0.785, which suggests that genetic factors explain a substantial portion of the variation in the binary trait (Gold), adjusted for the residual variance in the logit model.



