---
title: "everything5"
output: html_document
date: "2025-08-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### 1. Proportion

#startswith include gold overtone e.g gold silver, gold white

```{r}
dat <- PearlOperationalData

# Convert Colour to character (avoids factor issues)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove oysters with no pearls (missing or blank)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary Gold variable
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Convert Saibo donor ID to factor
dat$First_Op_Data_Saibo_No <- as.factor(dat$First_Op_Data_Saibo_No)

# Fit logistic regression
model <- glm(Gold ~ First_Op_Data_Saibo_No, data = dat, family = binomial)
summary(model)

```


#gold + gold overtones
```{r}
library(ggplot2)
library(dplyr)

# Calculate proportion of gold pearls per donor
prop_data <- dat %>%
  group_by(First_Op_Data_Saibo_No) %>%
  summarise(
    n = n(),
    gold_rate = mean(Gold)
  )

# Bar plot
p_gold <- ggplot(prop_data, aes(x = First_Op_Data_Saibo_No, y = gold_rate)) +
  geom_col(fill = "goldenrod") +
  labs(
    x = "Saibo Donor ID",
    y = "Proportion of Gold Pearls",
    title = "Proportion of Gold Pearls by Saibo Donor"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_gold

```


###silver + overtones silver gold, silver pink, silver white

```{r}
dat <- PearlOperationalData

# Convert Colour to character (avoids factor issues)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove oysters with no pearls (missing or blank)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary Gold variable
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Convert Saibo donor ID to factor
dat$First_Op_Data_Saibo_No <- as.factor(dat$First_Op_Data_Saibo_No)

# Fit logistic regression
model <- glm(Silver ~ First_Op_Data_Saibo_No, data = dat, family = binomial)
summary(model)

```


#silver + overtones

```{r}
library(ggplot2)
library(dplyr)

# Calculate proportion of gold pearls per donor
prop_data <- dat %>%
  group_by(First_Op_Data_Saibo_No) %>%
  summarise(
    n = n(),
    silver_rate = mean(Silver)
  )

# Bar plot
p_silver <- ggplot(prop_data, aes(x = First_Op_Data_Saibo_No, y = silver_rate)) +
  geom_col(fill = "#c0c0c0") +
  labs(
    x = "Saibo Donor ID",
    y = "Proportion of Silver Pearls",
    title = "Proportion of Silver Pearls by Saibo Donor"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_silver

```



# white + overtones white cream, white gold, white pink, white silver

```{r}
dat <- PearlOperationalData

# Convert Colour to character (avoids factor issues)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove oysters with no pearls (missing or blank)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary Gold variable
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Convert Saibo donor ID to factor
dat$First_Op_Data_Saibo_No <- as.factor(dat$First_Op_Data_Saibo_No)

# Fit logistic regression
model <- glm(White ~ First_Op_Data_Saibo_No, data = dat, family = binomial)
summary(model)

```

#white + overtones

```{r}
library(ggplot2)
library(dplyr)

# Calculate proportion of gold pearls per donor
prop_data <- dat %>%
  group_by(First_Op_Data_Saibo_No) %>%
  summarise(
    n = n(),
    white_rate = mean(White)
  )

# Bar plot
p_white <- ggplot(prop_data, aes(x = First_Op_Data_Saibo_No, y = white_rate)) +
  geom_col(fill = "#E3DAC9") +
  labs(
    x = "Saibo Donor ID",
    y = "Proportion of White Pearls",
    title = "Proportion of White Pearls by Saibo Donor"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_white

```



# cream 

```{r}
dat <- PearlOperationalData

# Convert Colour to character (avoids factor issues)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove oysters with no pearls (missing or blank)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary Gold variable
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Convert Saibo donor ID to factor
dat$First_Op_Data_Saibo_No <- as.factor(dat$First_Op_Data_Saibo_No)

# Fit logistic regression
model <- glm(Cream ~ First_Op_Data_Saibo_No, data = dat, family = binomial)
summary(model)

```


#cream 

```{r}
library(ggplot2)
library(dplyr)

# Calculate proportion of cream pearls per donor
prop_data <- dat %>%
  group_by(First_Op_Data_Saibo_No) %>%
  summarise(
    n = n(),
    cream_rate = mean(Cream)
  )

# Bar plot
p_cream <- ggplot(prop_data, aes(x = First_Op_Data_Saibo_No, y = cream_rate)) +
  geom_col(fill = "#FFFDD0") +
  labs(
    x = "Saibo Donor ID",
    y = "Proportion of Cream Pearls",
    title = "Proportion of Cream Pearls by Saibo Donor"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_cream

```



# champagne

```{r}
dat <- PearlOperationalData

# Convert Colour to character (avoids factor issues)
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove oysters with no pearls (missing or blank)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary Gold variable
dat$Champ <- ifelse(grepl("Champ",dat$Pearl_Grading_Data_Colour), 1, 0)

# Convert Saibo donor ID to factor
dat$First_Op_Data_Saibo_No <- as.factor(dat$First_Op_Data_Saibo_No)

# Fit logistic regression
model <- glm(Champ ~ First_Op_Data_Saibo_No, data = dat, family = binomial)
summary(model)

```


#champagne and light champ

```{r}
library(ggplot2)
library(dplyr)

# Calculate proportion of champ pearls per donor
prop_data <- dat %>%
  group_by(First_Op_Data_Saibo_No) %>%
  summarise(
    n = n(),
    champ_rate = mean(Champ)
  )

# Bar plot
p_champ <- ggplot(prop_data, aes(x = First_Op_Data_Saibo_No, y = champ_rate)) +
  geom_col(fill = "#F7E7CE") +
  labs(
    x = "Saibo Donor ID",
    y = "Proportion of Champagne Pearls",
    title = "Proportion of Champagne Pearls by Saibo Donor"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p_champ

```

#stacked plot

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)

# Start with your dataset
dat <- PearlOperationalData

# Clean and classify base colours (includes overtones)
dat <- dat %>%
  filter(!is.na(Pearl_Grading_Data_Colour),
         Pearl_Grading_Data_Colour != "") %>%
  mutate(
    base_colour = case_when(
      str_detect(Pearl_Grading_Data_Colour, regex("^Gold", ignore_case = TRUE)) ~ "Gold",
      str_detect(Pearl_Grading_Data_Colour, regex("^Silver", ignore_case = TRUE)) ~ "Silver",
      str_detect(Pearl_Grading_Data_Colour, regex("^White", ignore_case = TRUE)) ~ "White",
      str_detect(Pearl_Grading_Data_Colour, regex("^Cream", ignore_case = TRUE)) ~ "Cream",
      str_detect(Pearl_Grading_Data_Colour, regex("Champ", ignore_case = TRUE)) ~ "Champagne",
      TRUE ~ NA_character_
    ),
    First_Op_Data_Saibo_No = as.factor(First_Op_Data_Saibo_No)
  ) %>%
  filter(!is.na(base_colour))

# Ensure legend order
dat$base_colour <- factor(dat$base_colour,
                          levels = c("Gold", "Silver", "White", "Cream", "Champagne"))

# Count and calculate proportions per donor
stack_df <- dat %>%
  count(First_Op_Data_Saibo_No, base_colour, name = "n") %>%
  group_by(First_Op_Data_Saibo_No) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

# 100% stacked bar plot with custom colours
ggplot(stack_df, aes(x = First_Op_Data_Saibo_No, y = prop, fill = base_colour)) +
  geom_col(width = 0.9) +
  scale_fill_manual(
    values = c(
      "Gold" = "#ffd700",
      "Silver" = "#696969",
      "White" = "#d3d3d3",
      "Cream" = "#fffacd",
      "Champagne" = "#ffa07a"
    )
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    x = "Saibo Donor ID",
    y = "Pearl Colour Composition (%)",
    fill = "Base colour",
    title = "Pearl colour composition per Saibo donor (startswith grouping)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right"
  )

```



# 2. Fixed Effects

#First_Op_Data_DVH_mm
# gold + overtones
#not signif
```{r}
# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Gold vs Others
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Remove rows with missing First_Op_Data_DVH_mm
dat <- dat[!is.na(dat$First_Op_Data_DVH_mm), ]

# Fit logistic regression: DVH vs Gold
model <- glm(Gold ~ First_Op_Data_DVH_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```

#First_Op_Data_DVH_mm
#silver + overtones
#signif intercept// not signif
```{r}
# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Silver vs Others
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Remove rows with missing First_Op_Data_DVH_mm
dat <- dat[!is.na(dat$First_Op_Data_DVH_mm), ]

# Fit logistic regression: DVH vs Silver
model <- glm(Silver ~ First_Op_Data_DVH_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```


#First_Op_Data_DVH_mm
#white + overtones
#not signif

```{r}
# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: White vs Others
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Remove rows with missing First_Op_Data_DVH_mm
dat <- dat[!is.na(dat$First_Op_Data_DVH_mm), ]

# Fit logistic regression: DVH vs White
model <- glm(White ~ First_Op_Data_DVH_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```


#First_Op_Data_DVH_mm
#cream
#signif

```{r}
# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Cream vs Others
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Remove rows with missing First_Op_Data_DVH_mm
dat <- dat[!is.na(dat$First_Op_Data_DVH_mm), ]

# Fit logistic regression: DVH vs White
model <- glm(Cream ~ First_Op_Data_DVH_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```


#First_Op_Data_DVH_mm
#Champagne
#not signif

```{r}
# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Champ vs Others
dat$Champ <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# Remove rows with missing First_Op_Data_DVH_mm
dat <- dat[!is.na(dat$First_Op_Data_DVH_mm), ]

# Fit logistic regression: DVH vs White
model <- glm(Champ ~ First_Op_Data_DVH_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```



# First_Op_Data_APW_mm
#gold + overtones
#not signif
```{r}

# Load dataset
dat <- PearlOperationalData

# Convert Colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove missing/blank colour values
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary variable for Gold pearls
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Remove rows with missing APW values
dat <- dat[!is.na(dat$First_Op_Data_APW_mm), ]

# Fit logistic regression model
model <- glm(Gold ~ First_Op_Data_APW_mm, data = dat, family = binomial)

# Show model summary
summary(model)



```



# First_Op_Data_APW_mm
#silver + overtones
#signif intercept // not signif
```{r}

# Load dataset
dat <- PearlOperationalData

# Convert Colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove missing/blank colour values
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary variable for Silver pearls
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Remove rows with missing APW values
dat <- dat[!is.na(dat$First_Op_Data_APW_mm), ]

# Fit logistic regression model
model <- glm(Silver ~ First_Op_Data_APW_mm, data = dat, family = binomial)

# Show model summary
summary(model)



```

# First_Op_Data_APW_mm
#white + overtones
#not signif
```{r}

# Load dataset
dat <- PearlOperationalData

# Convert Colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove missing/blank colour values
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary variable for White pearls
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Remove rows with missing APW values
dat <- dat[!is.na(dat$First_Op_Data_APW_mm), ]

# Fit logistic regression model
model <- glm(White ~ First_Op_Data_APW_mm, data = dat, family = binomial)

# Show model summary
summary(model)



```

# First_Op_Data_APW_mm
#cream
#not signif
```{r}

# Load dataset
dat <- PearlOperationalData

# Convert Colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove missing/blank colour values
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary variable for Cream pearls
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Remove rows with missing APW values
dat <- dat[!is.na(dat$First_Op_Data_APW_mm), ]

# Fit logistic regression model
model <- glm(Cream ~ First_Op_Data_APW_mm, data = dat, family = binomial)

# Show model summary
summary(model)



```

# First_Op_Data_APW_mm
#champ
# not signif
```{r}

# Load dataset
dat <- PearlOperationalData

# Convert Colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Remove missing/blank colour values
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary variable for Champ pearls
dat$Champ <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# Remove rows with missing APW values
dat <- dat[!is.na(dat$First_Op_Data_APW_mm), ]

# Fit logistic regression model
model <- glm(Champ ~ First_Op_Data_APW_mm, data = dat, family = binomial)

# Show model summary
summary(model)



```


## 

##gold 
#  First_Op_Data_Shell_Depth_mm
# not signif

```{r}

# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Gold vs Others
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Depth_mm
dat <- dat[!is.na(dat$First_Op_Data_Shell_Depth_mm), ]

# Fit logistic regression: DVH vs Gold
model <- glm(Gold ~ First_Op_Data_Shell_Depth_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```

#silver 
#First_Op_Data_Shell_Depth_mm
#not signif


```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Silver vs Others
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Depth_mm
dat <- dat[!is.na(dat$First_Op_Data_Shell_Depth_mm), ]

# Fit logistic regression: DVH vs Silver
model <- glm(Silver ~ First_Op_Data_Shell_Depth_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```

#white
#First_Op_Data_Shell_Depth_mm
#not signif
```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: White vs Others
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Depth_mm
dat <- dat[!is.na(dat$First_Op_Data_Shell_Depth_mm), ]

# Fit logistic regression: DVH vs White
model <- glm(White ~ First_Op_Data_Shell_Depth_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```

#cream
# First_Op_Data_Shell_Depth_mm
#not signif


```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Cream vs Others
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Depth_mm
dat <- dat[!is.na(dat$First_Op_Data_Shell_Depth_mm), ]

# Fit logistic regression: DVH vs Cream
model <- glm(Cream ~ First_Op_Data_Shell_Depth_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```


#champ
#First_Op_Data_Shell_Depth_mm
#not signif
```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Champ vs Others
dat$Champ <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Depth_mm
dat <- dat[!is.na(dat$First_Op_Data_Shell_Depth_mm), ]

# Fit logistic regression: DVH vs Cream
model <- glm(Champ ~ First_Op_Data_Shell_Depth_mm, data = dat, family = binomial)

# Summary of model
summary(model)

```

## statistical test for shell weight vs binary colour 

##gold 
#First_Op_Data_Shell_Weight_g
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Gold vs Others
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Weight_g
dat <- dat[!is.na(dat$First_Op_Data_Shell_Weight_g), ]

# Fit logistic regression: shell weight vs Gold
model <- glm(Gold ~ First_Op_Data_Shell_Weight_g, data = dat, family = binomial)

# Summary of model
summary(model)

```


##silver
#First_Op_Data_Shell_Weight_g
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Silver vs Others
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Weight_g
dat <- dat[!is.na(dat$First_Op_Data_Shell_Weight_g), ]

# Fit logistic regression: shell weight vs Gold
model <- glm(Silver ~ First_Op_Data_Shell_Weight_g, data = dat, family = binomial)

# Summary of model
summary(model)

```

##white
#First_Op_Data_Shell_Weight_g
#not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: White vs Others
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Weight_g
dat <- dat[!is.na(dat$First_Op_Data_Shell_Weight_g), ]

# Fit logistic regression: shell weight vs Gold
model <- glm(White ~ First_Op_Data_Shell_Weight_g, data = dat, family = binomial)

# Summary of model
summary(model)

```


##cream
#First_Op_Data_Shell_Weight_g
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Cream vs Others
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Weight_g
dat <- dat[!is.na(dat$First_Op_Data_Shell_Weight_g), ]

# Fit logistic regression: shell weight vs Cream
model <- glm(Cream ~ First_Op_Data_Shell_Weight_g, data = dat, family = binomial)

# Summary of model
summary(model)

```

##champagne
#First_Op_Data_Shell_Weight_g
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: Champ vs Others
dat$Champ <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# Remove rows with missing First_Op_Data_Shell_Weight_g
dat <- dat[!is.na(dat$First_Op_Data_Shell_Weight_g), ]

# Fit logistic regression: shell weight vs Cream
model <- glm(Champ ~ First_Op_Data_Shell_Weight_g, data = dat, family = binomial)

# Summary of model
summary(model)

```




#gold
#First_Op_Data_Nacre_Lip_Colour_Round
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: gold vs Others
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)

# Remove rows with missing First_Op_Data_Nacre_Lip_Colour_Round
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Round), ]

# Fit logistic regression: shell weight vs gold
model <- glm(Gold ~ First_Op_Data_Nacre_Lip_Colour_Round, data = dat, family = binomial)

# Summary of model
summary(model)

```



#silver
#First_Op_Data_Nacre_Lip_Colour_Round
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: silver vs Others
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)

# Remove rows with missing First_Op_Data_Nacre_Lip_Colour_Round
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Round), ]

# Fit logistic regression: nacre lip colour host vs Silver
model <- glm(Silver ~ First_Op_Data_Nacre_Lip_Colour_Round, data = dat, family = binomial)

# Summary of model
summary(model)

```



#white
#First_Op_Data_Nacre_Lip_Colour_Round
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: silver vs Others
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)

# Remove rows with missing First_Op_Data_Nacre_Lip_Colour_Round
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Round), ]

# Fit logistic regression: nacre lip colour host vs Silver
model <- glm(White ~ First_Op_Data_Nacre_Lip_Colour_Round, data = dat, family = binomial)

# Summary of model
summary(model)

```

#cream
#First_Op_Data_Nacre_Lip_Colour_Round
#intercept signif// not signif

```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: cream vs Others
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)

# Remove rows with missing First_Op_Data_Nacre_Lip_Colour_Round
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Round), ]

# Fit logistic regression: nacre lip colour host vs cream
model <- glm(Cream ~ First_Op_Data_Nacre_Lip_Colour_Round, data = dat, family = binomial)

# Summary of model
summary(model)

```

#champ
#First_Op_Data_Nacre_Lip_Colour_Round
#intercept signif// not signif


```{r}



# Load your dataset
dat <- PearlOperationalData

# Convert colour to character
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)

# Filter out blanks/missing
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]

# Create binary: champ vs Others
dat$Champ <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# Remove rows with missing First_Op_Data_Nacre_Lip_Colour_Round
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Round), ]

# Fit logistic regression: nacre lip colour host vs champ
model <- glm(Champ ~ First_Op_Data_Nacre_Lip_Colour_Round, data = dat, family = binomial)

# Summary of model
summary(model)

```

#gold, silver, white, cream, champ
#First_Op_Data_Nacre_Lip_Colour_Flat
#intercept signif// not signif for all

```{r}
library(ggplot2)
library(dplyr)

# Load and clean data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$First_Op_Data_Nacre_Lip_Colour_Flat), ]

# Convert to factor
dat$First_Op_Data_Nacre_Lip_Colour_Flat <- factor(dat$First_Op_Data_Nacre_Lip_Colour_Flat)

# Create binary columns for each pearl colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Fit logistic models
mod_gold <- glm(Gold ~ First_Op_Data_Nacre_Lip_Colour_Flat, data = dat, family = binomial)
mod_silver <- glm(Silver ~ First_Op_Data_Nacre_Lip_Colour_Flat, data = dat, family = binomial)
mod_white <- glm(White ~ First_Op_Data_Nacre_Lip_Colour_Flat, data = dat, family = binomial)
mod_cream <- glm(Cream ~ First_Op_Data_Nacre_Lip_Colour_Flat, data = dat, family = binomial)
mod_champ <- glm(Champ ~ First_Op_Data_Nacre_Lip_Colour_Flat, data = dat, family = binomial)

# GLM statistical tests
summary(mod_gold)   # Test: Does lip colour predict Gold pearl?
summary(mod_silver) # Test: Does lip colour predict Silver pearl?
summary(mod_white)  # Test: Does lip colour predict White pearl?
summary(mod_cream)  # Test: Does lip colour predict Cream pearl?
summary(mod_champ)  # Test: Does lip colour predict Champ pearl?

```


#gold, silver, white, cream, champ
#First_Op_Data_Nuclei_Size_mm
#some intercept signif// silver and cream signif


```{r}
library(ggplot2)
library(dplyr)

# Load and clean data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$First_Op_Data_Nuclei_Size_mm), ]

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Fit logistic regression models
mod_gold <- glm(Gold ~ First_Op_Data_Nuclei_Size_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ First_Op_Data_Nuclei_Size_mm, data = dat, family = binomial)
mod_white <- glm(White ~ First_Op_Data_Nuclei_Size_mm, data = dat, family = binomial)
mod_cream <- glm(Cream ~ First_Op_Data_Nuclei_Size_mm, data = dat, family = binomial)
mod_champ <- glm(Champ ~ First_Op_Data_Nuclei_Size_mm, data = dat, family = binomial)

# GLM summaries
cat("==== Gold ====\n")
print(summary(mod_gold))       # Tests if nuclei size affects probability of gold pearl
print(anova(mod_gold, test = "Chisq"))  # Likelihood ratio test

cat("==== Silver ====\n")
print(summary(mod_silver))
print(anova(mod_silver, test = "Chisq"))

cat("==== White ====\n")
print(summary(mod_white))
print(anova(mod_white, test = "Chisq"))

cat("==== Cream ====\n")
print(summary(mod_cream))
print(anova(mod_cream, test = "Chisq"))

cat("==== Champ ====\n")
print(summary(mod_champ))
print(anova(mod_champ, test = "Chisq"))

```




#gold, silver, white, cream, champ
#First_Op_Data_Saibo_Area
#some intercept signif // silver and white signif

```{r}
library(ggplot2)
library(dplyr)

# Load and clean data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$First_Op_Data_Saibo_Area), ]
dat$First_Op_Data_Saibo_Area <- as.factor(dat$First_Op_Data_Saibo_Area)

# Create binary outcome variables for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Fit logistic regression models
mod_gold <- glm(Gold ~ First_Op_Data_Saibo_Area, data = dat, family = binomial)
mod_silver <- glm(Silver ~ First_Op_Data_Saibo_Area, data = dat, family = binomial)
mod_white <- glm(White ~ First_Op_Data_Saibo_Area, data = dat, family = binomial)
mod_cream <- glm(Cream ~ First_Op_Data_Saibo_Area, data = dat, family = binomial)
mod_champ <- glm(Champ ~ First_Op_Data_Saibo_Area, data = dat, family = binomial)

summary(glm(Gold ~ First_Op_Data_Saibo_Area, data = dat, family = binomial))
summary(glm(Silver ~ First_Op_Data_Saibo_Area, data = dat, family = binomial))
summary(glm(White ~ First_Op_Data_Saibo_Area, data = dat, family = binomial))
summary(glm(Cream ~ First_Op_Data_Saibo_Area, data = dat, family = binomial))
summary(glm(Champ ~ First_Op_Data_Saibo_Area, data = dat, family = binomial))



```




#Pearl_Harvest_Data_Days_of_Pearl_Culture
#champ signif
## statistical test
#Test if the proportion (rate) of a color changes with Days of Pearl Culture (continuous variable)
#Use logistic regression (generalized linear model, binomial family) for each color:

```{r}
# Example for Gold pearls: assuming Gold is binary 0/1 per pearl
glm_gold <- glm(Gold ~ Pearl_Harvest_Data_Days_of_Pearl_Culture, data = dat, family = binomial)
summary(glm_gold)
#This tests whether the odds of a pearl being Gold change with Days of Pearl Culture.

glm_silver <- glm(Silver ~ Pearl_Harvest_Data_Days_of_Pearl_Culture, data = dat, family = binomial)
summary(glm_silver)

glm_white <- glm(White ~ Pearl_Harvest_Data_Days_of_Pearl_Culture, data = dat, family = binomial)
summary(glm_white)

glm_cream <- glm(Cream ~ Pearl_Harvest_Data_Days_of_Pearl_Culture, data = dat, family = binomial)
summary(glm_cream)

glm_champ <- glm(Champ ~ Pearl_Harvest_Data_Days_of_Pearl_Culture, data = dat, family = binomial)
summary(glm_champ)


```


#gold, silver, white, cream, champ
#Pearl_Harvest_Data_DVH_mm
#some intercept signif// gold and white signif

```{r}
library(ggplot2)
library(dplyr)

# Load and prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Pearl_Harvest_Data_DVH_mm), ]

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)


# Logistic regression models per colour
mod_gold   <- glm(Gold ~ Pearl_Harvest_Data_DVH_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Pearl_Harvest_Data_DVH_mm, data = dat, family = binomial)
mod_white  <- glm(White ~ Pearl_Harvest_Data_DVH_mm, data = dat, family = binomial)
mod_cream  <- glm(Cream ~ Pearl_Harvest_Data_DVH_mm, data = dat, family = binomial)
mod_champ  <- glm(Champ ~ Pearl_Harvest_Data_DVH_mm, data = dat, family = binomial)

# Display GLM summaries (coefficients, p-values, etc.)
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)



```


#gold, silver, white, cream, champ
#Pearl_Harvest_Data_APW_mm
#gold signif

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Pearl_Harvest_Data_APW_mm), ]  # Use APW instead of DVH

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)


# Fit GLMs for each colour vs APW (logistic regression)
mod_gold   <- glm(Gold ~ Pearl_Harvest_Data_APW_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Pearl_Harvest_Data_APW_mm, data = dat, family = binomial)
mod_white  <- glm(White ~ Pearl_Harvest_Data_APW_mm, data = dat, family = binomial)
mod_cream  <- glm(Cream ~ Pearl_Harvest_Data_APW_mm, data = dat, family = binomial)
mod_champ  <- glm(Champ ~ Pearl_Harvest_Data_APW_mm, data = dat, family = binomial)

# Show GLM summaries (coefficients, p-values, etc.)
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)



```


#gold, silver, white, cream, champ
#Pearl_Harvest_Data_Shell_Depth_mm
#some intercept signif// not signif for all

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Pearl_Harvest_Data_Shell_Depth_mm), ]  # Use Shell Depth

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Logistic regression models
mod_gold   <- glm(Gold ~ Pearl_Harvest_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Pearl_Harvest_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_white  <- glm(White ~ Pearl_Harvest_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_cream  <- glm(Cream ~ Pearl_Harvest_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_champ  <- glm(Champ ~ Pearl_Harvest_Data_Shell_Depth_mm, data = dat, family = binomial)

# View model summaries
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)
```



#gold, silver, white, cream, champ
#Pearl_Harvest_Data_Shell_Weight_g
# some intercept signif// gold and white signif

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Pearl_Harvest_Data_Shell_Weight_g), ]  # Use Shell Weight

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Fit logistic regression models
mod_gold <- glm(Gold ~ Pearl_Harvest_Data_Shell_Weight_g, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Pearl_Harvest_Data_Shell_Weight_g, data = dat, family = binomial)
mod_white <- glm(White ~ Pearl_Harvest_Data_Shell_Weight_g, data = dat, family = binomial)
mod_cream <- glm(Cream ~ Pearl_Harvest_Data_Shell_Weight_g, data = dat, family = binomial)
mod_champ <- glm(Champ ~ Pearl_Harvest_Data_Shell_Weight_g, data = dat, family = binomial)

# Get summaries of each model (includes p-values)
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)


```


##donor
# gold, silver, white, cream, champ
# Matched_Donor_Data_DVH_mm
# some intercept signif // cream signif

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Matched_Donor_Data_DVH_mm), ]  # Use Donor DVH

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)


# Fit logistic regression models
mod_gold <- glm(Gold ~ Matched_Donor_Data_DVH_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_DVH_mm, data = dat, family = binomial)
mod_white <- glm(White ~ Matched_Donor_Data_DVH_mm, data = dat, family = binomial)
mod_cream <- glm(Cream ~ Matched_Donor_Data_DVH_mm, data = dat, family = binomial)
mod_champ <- glm(Champ ~ Matched_Donor_Data_DVH_mm, data = dat, family = binomial)

summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)


```



#gold, silver, white, cream, champ
#Matched_Donor_Data_APW_mm
#some intercept signif// not signif for all


```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Matched_Donor_Data_APW_mm), ]  # Use Donor APW

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)

# Fit GLMs
mod_gold   <- glm(Gold ~ Matched_Donor_Data_APW_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_APW_mm, data = dat, family = binomial)
mod_white  <- glm(White ~ Matched_Donor_Data_APW_mm, data = dat, family = binomial)
mod_cream  <- glm(Cream ~ Matched_Donor_Data_APW_mm, data = dat, family = binomial)
mod_champ  <- glm(Champ ~ Matched_Donor_Data_APW_mm, data = dat, family = binomial)

summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)
```


#gold, silver, white, cream, champ
#Matched_Donor_Data_Shell_Depth_mm
#some intercept signif// not signif for all

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Matched_Donor_Data_Shell_Depth_mm), ]  # Use Donor Shell Depth

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)


# Fit logistic models for each colour
mod_gold   <- glm(Gold ~ Matched_Donor_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_white  <- glm(White ~ Matched_Donor_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_cream  <- glm(Cream ~ Matched_Donor_Data_Shell_Depth_mm, data = dat, family = binomial)
mod_champ  <- glm(Champ ~ Matched_Donor_Data_Shell_Depth_mm, data = dat, family = binomial)



summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)

```



#gold, silver, white, cream, champ
#Matched_Donor_Data_Shell_Weight_g
#some intercept signif// cream signif

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Prepare data
dat <- PearlOperationalData
dat$Pearl_Grading_Data_Colour <- as.character(dat$Pearl_Grading_Data_Colour)
dat <- dat[!is.na(dat$Pearl_Grading_Data_Colour) & dat$Pearl_Grading_Data_Colour != "", ]
dat <- dat[!is.na(dat$Matched_Donor_Data_Shell_Weight_g), ]  # Use Donor Shell Weight

# Create binary columns for each colour
dat$Gold <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
dat$Silver <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
dat$White <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
dat$Cream <- ifelse(startsWith(dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
dat$Champ <- ifelse(grepl("Champ", dat$Pearl_Grading_Data_Colour), 1, 0)


# Fit logistic regression models
mod_gold   <- glm(Gold   ~ Matched_Donor_Data_Shell_Weight_g, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_Shell_Weight_g, data = dat, family = binomial)
mod_white  <- glm(White  ~ Matched_Donor_Data_Shell_Weight_g, data = dat, family = binomial)
mod_cream  <- glm(Cream  ~ Matched_Donor_Data_Shell_Weight_g, data = dat, family = binomial)
mod_champ  <- glm(Champ  ~ Matched_Donor_Data_Shell_Weight_g, data = dat, family = binomial)

summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)


```




#gold, silver, white, cream, champ
#Matched_Donor_Data_Nacre_lip_colour_Round
#those with Gr (green) is usually gold, cream // gold, white, cream, champ signif



```{r}
library(dplyr)
library(ggplot2)

# Load and clean data
dat <- PearlOperationalData

# Keep only rows with valid colour and lip colour
dat <- dat %>%
  filter(
    !is.na(Pearl_Grading_Data_Colour),
    Pearl_Grading_Data_Colour != "",
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Round != ""
  )

# Merge rare lip colour levels into "Other"
dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.character(dat$Matched_Donor_Data_Nacre_lip_colour_Round)
dat$Matched_Donor_Data_Nacre_lip_colour_Round <- ifelse(
  dat$Matched_Donor_Data_Nacre_lip_colour_Round %in% c("WY", "LGr"),
  "Other",
  dat$Matched_Donor_Data_Nacre_lip_colour_Round
)

# Convert to factor and set reference level to "S"
dat$Matched_Donor_Data_Nacre_lip_colour_Round <- factor(dat$Matched_Donor_Data_Nacre_lip_colour_Round)
if ("S" %in% levels(dat$Matched_Donor_Data_Nacre_lip_colour_Round)) {
  dat$Matched_Donor_Data_Nacre_lip_colour_Round <- relevel(dat$Matched_Donor_Data_Nacre_lip_colour_Round, ref = "S")
}

# View updated distribution of lip colours
print(table(dat$Matched_Donor_Data_Nacre_lip_colour_Round))

# Create binary outcome variables for each colour (case-insensitive)
is_start <- function(x, prefix) grepl(paste0("^", prefix), x, ignore.case = TRUE)
dat$Gold   <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Gold"))
dat$Silver <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Silver"))
dat$White  <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "White"))
dat$Cream  <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Cream"))
dat$Champ  <- as.integer(grepl("Champ", dat$Pearl_Grading_Data_Colour, ignore.case = TRUE))

# Fit logistic regression models
mod_gold   <- glm(Gold   ~ Matched_Donor_Data_Nacre_lip_colour_Round, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_Nacre_lip_colour_Round, data = dat, family = binomial)
mod_white  <- glm(White  ~ Matched_Donor_Data_Nacre_lip_colour_Round, data = dat, family = binomial)
mod_cream  <- glm(Cream  ~ Matched_Donor_Data_Nacre_lip_colour_Round, data = dat, family = binomial)
mod_champ  <- glm(Champ  ~ Matched_Donor_Data_Nacre_lip_colour_Round, data = dat, family = binomial)

# Summarise models
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)

```




#gold, silver, white, cream, champ
#Matched_Donor_Data_Nacre_lip_colour_Flat
#some intercepts are signif// gold, white, cream, champ signif; Silver and White do not have Gr


```{r}
library(dplyr)
library(ggplot2)

# Load and clean data
dat <- PearlOperationalData

# Keep only rows with valid pearl colour and flat lip colour
dat <- dat %>%
  filter(
    !is.na(Pearl_Grading_Data_Colour),
    Pearl_Grading_Data_Colour != "",
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    Matched_Donor_Data_Nacre_lip_colour_Flat != ""
  )

# Merge rare flat lip colour levels into "Other"
dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.character(dat$Matched_Donor_Data_Nacre_lip_colour_Flat)
dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- ifelse(
  dat$Matched_Donor_Data_Nacre_lip_colour_Flat %in% c("WY", "LGr"),
  "Other",
  dat$Matched_Donor_Data_Nacre_lip_colour_Flat
)

# Convert to factor and set reference level to "S"
dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- factor(dat$Matched_Donor_Data_Nacre_lip_colour_Flat)
if ("S" %in% levels(dat$Matched_Donor_Data_Nacre_lip_colour_Flat)) {
  dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- relevel(dat$Matched_Donor_Data_Nacre_lip_colour_Flat, ref = "S")
}

# View updated distribution of lip colours
print(table(dat$Matched_Donor_Data_Nacre_lip_colour_Flat))

# Create binary outcome variables for each colour (case-insensitive)
is_start <- function(x, prefix) grepl(paste0("^", prefix), x, ignore.case = TRUE)
dat$Gold   <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Gold"))
dat$Silver <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Silver"))
dat$White  <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "White"))
dat$Cream  <- as.integer(is_start(dat$Pearl_Grading_Data_Colour, "Cream"))
dat$Champ  <- as.integer(grepl("Champ", dat$Pearl_Grading_Data_Colour, ignore.case = TRUE))

# Fit logistic regression models
mod_gold   <- glm(Gold   ~ Matched_Donor_Data_Nacre_lip_colour_Flat, data = dat, family = binomial)
mod_silver <- glm(Silver ~ Matched_Donor_Data_Nacre_lip_colour_Flat, data = dat, family = binomial)
mod_white  <- glm(White  ~ Matched_Donor_Data_Nacre_lip_colour_Flat, data = dat, family = binomial)
mod_cream  <- glm(Cream  ~ Matched_Donor_Data_Nacre_lip_colour_Flat, data = dat, family = binomial)
mod_champ  <- glm(Champ  ~ Matched_Donor_Data_Nacre_lip_colour_Flat, data = dat, family = binomial)

# Summarise models
summary(mod_gold)
summary(mod_silver)
summary(mod_white)
summary(mod_cream)
summary(mod_champ)

```



# 3. Traditional Model (Binary)

## TRADITIONAL HERITABILITY ESTIMATES AND ESTIMATED BREEDING VALUES


```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Gold),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Gold ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Gold)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```

# BINARY MODEL SILVER
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Silver),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Silver ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Silver)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```





# BINARY MODEL WHITE
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(White),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = White ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$White)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```


# BINARY MODEL CREAM
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Cream),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Cream ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Cream)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```



# BINARY MODEL CHAMPAGNE
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Champ),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Champ ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Champ)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```



#DONOR 
GOLD BINARY MODEL #with fixed effects

```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Gold),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#pearson correlation binary model gold

```{r}
# Load required library
library(ggplot2)

# --- Step 1: Rename EBV column and ensure donor_id is a factor ---
BLUP <- BLUP %>%
  rename(EBV = solution) %>%
  mutate(donor_id = as.factor(donor_id))


# --- Step 2: Calculate observed proportion of Gold pearls per donor ---
obs_gold <- donor_dat_clean %>%
  group_by(donor_id) %>%
  summarise(
    prop_gold = mean(Gold, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Step 3: Merge EBVs with observed gold proportion ---
plot_df <- BLUP %>%
  inner_join(obs_gold, by = "donor_id") %>%
  filter(!is.na(EBV), !is.na(prop_gold))

# --- Step 4: Exponential model fit ---
plot_df <- plot_df %>%
  mutate(prop_gold_adj = pmax(prop_gold, 1e-6))  # avoid log(0)

# Starting values from log-linear regression
start_vals <- coef(lm(log(prop_gold_adj) ~ EBV, data = plot_df))
start_a <- exp(start_vals[1])
start_b <- start_vals[2]

# Fit exponential model: y = a * exp(b * x)
exp_model <- nls(prop_gold_adj ~ a * exp(b * EBV),
                 data = plot_df,
                 start = list(a = start_a, b = start_b),
                 control = list(maxiter = 200, warnOnly = TRUE))

# --- Step 5: Generate fitted curve ---
curve_df <- data.frame(EBV = seq(min(plot_df$EBV), max(plot_df$EBV), length.out = 200))
curve_df$pred <- predict(exp_model, newdata = curve_df)

# --- Step 6: Plot without point size ---
ggplot(plot_df, aes(x = EBV, y = prop_gold)) +
  geom_point(color = "darkgoldenrod", alpha = 0.8) +
  geom_line(data = curve_df, aes(x = EBV, y = pred), color = "black", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    x = "Estimated Breeding Value (EBV) for Gold",
    y = "Observed Proportion of Gold Pearls",
    title = "Donor EBV vs Observed Gold Proportion",
    subtitle = "Fitted with exponential model: y = a  exp(b  EBV)"
  ) +
  theme_minimal(base_size = 13)

# Optional: print model coefficients
cat("Fitted exponential model:\n")
print(coef(exp_model))

cor(plot_df$EBV, plot_df$prop_gold, method = "pearson", use = "complete.obs")

cor.test(plot_df$EBV, plot_df$prop_gold, method = "pearson")



```



DONOR BINARY SILVER NO FIXED EFFECTS
```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- read_csv("donor_data2.csv")


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Silver),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Silver ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Silver, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#DONOR 
WHITE BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(White),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = White ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$White, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#DONOR 
CREAM BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)
donor_dat$Matched_Donor_Data_DVH_mm <- as.numeric(donor_dat$Matched_Donor_Data_DVH_mm)
donor_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(donor_dat$Matched_Donor_Data_Shell_Weight_g)



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Cream),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Cream ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Cream, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#DONOR 
CHAMP BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)




# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Champ),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Champ ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Champ, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```




#vpredict quantitative model host 
```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- hostquant %>%
  filter(!is.na(offspring_id_host) & 
         !is.na(dam_host) & 
         !is.na(sire_host) & 
         !is.na(pred))

# Convert ID columns to factor
host_dat <- host_dat %>%
  mutate(
    offspring_id_host = as.factor(offspring_id_host),
    dam_host = as.factor(dam_host),
    sire_host = as.factor(sire_host),
    
    # Fixed effects
    First_Op_Data_DVH_mm = as.numeric(First_Op_Data_DVH_mm),
    First_Op_Data_Nuclei_Size_mm = as.numeric(First_Op_Data_Nuclei_Size_mm),
    First_Op_Data_Saibo_Area = as.factor(First_Op_Data_Saibo_Area),
    Pearl_Harvest_Data_Days_of_Pearl_Culture = as.factor(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    Pearl_Harvest_Data_DVH_mm = as.numeric(Pearl_Harvest_Data_DVH_mm),
    Pearl_Harvest_Data_APW_mm = as.numeric(Pearl_Harvest_Data_APW_mm),
    Pearl_Harvest_Data_Shell_Weight_g = as.numeric(Pearl_Harvest_Data_Shell_Weight_g),
    Matched_Donor_Data_DVH_mm = as.numeric(Matched_Donor_Data_DVH_mm),
    Matched_Donor_Data_Shell_Weight_g = as.numeric(Matched_Donor_Data_Shell_Weight_g),
    Matched_Donor_Data_Nacre_lip_colour_Round = as.factor(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Flat  = as.factor(Matched_Donor_Data_Nacre_lip_colour_Flat)
  )

# Drop unused factor levels
host_dat <- droplevels(host_dat)

# Filter complete cases for all model variables
host_dat_clean <- host_dat %>%
  filter(complete.cases(
    pred,
    offspring_id_host,
    First_Op_Data_DVH_mm,
    First_Op_Data_Nuclei_Size_mm,
    First_Op_Data_Saibo_Area,
    Pearl_Harvest_Data_Days_of_Pearl_Culture,
    Pearl_Harvest_Data_DVH_mm,
    Pearl_Harvest_Data_APW_mm,
    Pearl_Harvest_Data_Shell_Weight_g,
    Matched_Donor_Data_DVH_mm,
    Matched_Donor_Data_Shell_Weight_g,
    Matched_Donor_Data_Nacre_lip_colour_Round,
    Matched_Donor_Data_Nacre_lip_colour_Flat
  ))

# Add 'units' column
host_dat_clean$units <- 1:nrow(host_dat_clean)

# Prepare pedigree (no distinct used)
ped <- host_dat_clean %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Fit ASReml Gaussian animal model with fixed effects
mod_host <- asreml(
  fixed = pred ~ First_Op_Data_DVH_mm +
                 First_Op_Data_Nuclei_Size_mm +
                 First_Op_Data_Saibo_Area +
                 Pearl_Harvest_Data_Days_of_Pearl_Culture +
                 Pearl_Harvest_Data_DVH_mm +
                 Pearl_Harvest_Data_APW_mm +
                 Pearl_Harvest_Data_Shell_Weight_g +
                 Matched_Donor_Data_DVH_mm +
                 Matched_Donor_Data_Shell_Weight_g +
                 Matched_Donor_Data_Nacre_lip_colour_Round +
                 Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)

# Update model to ensure convergence
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)

# Extract and print variance components
vc <- summary(mod_host)$varcomp
print(vc)

# Heritability (latent, via vpredict)
vpredict(mod_host, h2 ~ V1 / (V1 + V2))

# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>%
  filter(host %in% unique(host_dat_clean$offspring_id_host)) %>%
  mutate(EBV = solution)

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```



```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- donorquant %>%
  filter(!is.na(offspring_id_donor) & 
         !is.na(dam_donor) & 
         !is.na(sire_donor) & 
         !is.na(pred))

# Convert columns to appropriate types
donor_dat <- donor_dat %>%
  mutate(
    offspring_id_donor = as.factor(offspring_id_donor),
    dam_donor = as.factor(dam_donor),
    sire_donor = as.factor(sire_donor),
    Matched_Donor_Data_Nacre_lip_colour_Round = as.factor(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Flat  = as.factor(Matched_Donor_Data_Nacre_lip_colour_Flat)
  )

# Drop unused factor levels
donor_dat <- droplevels(donor_dat)

# Filter complete cases including new fixed effects
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(
    pred,
    offspring_id_donor,
    Matched_Donor_Data_Nacre_lip_colour_Round,
    Matched_Donor_Data_Nacre_lip_colour_Flat
  ))

# Add units column (required by ASReml)
donor_dat_clean$units <- 1:nrow(donor_dat_clean)

# Prepare pedigree
ped <- donor_dat_clean %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Fit ASReml Gaussian animal model with the two fixed effects
mod_gold <- asreml(
  fixed = pred ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components
vc <- summary(mod_gold)$varcomp
print(vc)
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Calculate heritability
vpredict(mod_gold, h2 ~ V1 / (V1 + V2))

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>%
  filter(donor %in% unique(donor_dat_clean$offspring_id_donor)) %>%
  rename(EBV = solution)

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)

```



# 5. No. of Offspring per Donor Family

#NEW HOST FAMILY DISTRIBUTION WITH COLOUR #with NA

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Create family_key and family_number
pedigree_host <- pedigree_host2 %>%
  mutate(
    family_key = paste(sire_host, dam_host, sep = "_"),
    family_number = as.integer(factor(family_key))
  )

# Step 2: Create binary colour columns (edit "Champ" to "Champagne" for consistency)
pedigree_host <- pedigree_host %>%
  mutate(
    Gold       = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver     = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White      = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream      = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champagne  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)  # renamed
  )

# Step 3: Summarise by family_number
family_summary <- pedigree_host %>%
  group_by(family_number) %>%
  summarise(
    family_size    = n(),
    Gold_count     = sum(Gold, na.rm = TRUE),
    Silver_count   = sum(Silver, na.rm = TRUE),
    White_count    = sum(White, na.rm = TRUE),
    Cream_count    = sum(Cream, na.rm = TRUE),
    Champagne_count = sum(Champagne, na.rm = TRUE)
  ) %>%
  arrange(desc(family_size))  # Optional sorting

# Step 4: Reshape to long format for ggplot
family_long <- family_summary %>%
  pivot_longer(cols = c(Gold_count, Silver_count, White_count, Cream_count, Champagne_count),
               names_to = "Colour",
               values_to = "Count") %>%
  mutate(Colour = gsub("_count", "", Colour))  # Remove "_count" suffix

# Step 5: Define custom colours
my_colours <- c(
  Gold = "#FFD700",
  Silver = "#A9A9A9",
  White = "#D3D3D3",
  Cream = "#FFFDD0",
  Champagne = "#F7E7CE"
)

# Step 6: Create stacked bar plot
ggplot(family_long, aes(x = factor(family_number), y = Count, fill = Colour)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Family Number",
    y = "Pearl Count",
    title = "Distribution of Pearl Colours Across Families"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = my_colours)



```


#without NA

#GOOD distribution of pearl colours across families host
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Create family_key and family_number
pedigree_host <- pedigree_host2 %>%
  mutate(
    family_key = paste(sire_host, dam_host, sep = "_"),
    family_number = as.integer(factor(family_key))
  )

# Step 2: Create binary colour columns
pedigree_host <- pedigree_host %>%
  mutate(
    Gold       = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver     = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White      = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream      = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champagne  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# Step 3: Summarise by family_number
family_summary <- pedigree_host %>%
  group_by(family_number) %>%
  summarise(
    family_size     = n(),
    Gold_count      = sum(Gold, na.rm = TRUE),
    Silver_count    = sum(Silver, na.rm = TRUE),
    White_count     = sum(White, na.rm = TRUE),
    Cream_count     = sum(Cream, na.rm = TRUE),
    Champagne_count = sum(Champagne, na.rm = TRUE)
  ) %>%
  arrange(desc(family_size))

# Step 4: Reshape and remove NA or 0 counts
family_long <- family_summary %>%
  pivot_longer(cols = c(Gold_count, Silver_count, White_count, Cream_count, Champagne_count),
               names_to = "Colour",
               values_to = "Count") %>%
  mutate(Colour = gsub("_count", "", Colour)) %>%
  filter(!is.na(Count) & Count > 0)  # Remove NA and 0s

# Step 5: Define custom colours
my_colours <- c(
  Gold = "#FFD700",
  Silver = "#A9A9A9",
  White = "#D3D3D3",
  Cream = "#FFFDD0",
  Champagne = "#F7E7CE"
)

# Step 6: Plot
ggplot(family_long, aes(x = factor(family_number), y = Count, fill = Colour)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Family Number",
    y = "Pearl Count",
    title = "Distribution of Pearl Colours Across Families"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = my_colours)

```




#NEW DONOR 

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)

# Step 1: Create family_key and family_number
pedigree_donor <- donor_data2 %>%
  mutate(
    family_key = paste(`Candidate father ID`, `Candidate mother ID`, sep = "_"),
    family_number = as.integer(factor(family_key))
  )

# Step 2: Create binary colour columns
pedigree_donor <- pedigree_donor %>%
  mutate(
    Gold       = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver     = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White      = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream      = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champagne  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# Step 3: Summarise by family_number
family_summary <- pedigree_donor %>%
  group_by(family_number) %>%
  summarise(
    family_size     = n(),
    Gold_count      = sum(Gold, na.rm = TRUE),
    Silver_count    = sum(Silver, na.rm = TRUE),
    White_count     = sum(White, na.rm = TRUE),
    Cream_count     = sum(Cream, na.rm = TRUE),
    Champagne_count = sum(Champagne, na.rm = TRUE)
  ) %>%
  arrange(desc(family_size))

# Step 4: Reshape and remove NA or 0 counts
family_long <- family_summary %>%
  pivot_longer(cols = c(Gold_count, Silver_count, White_count, Cream_count, Champagne_count),
               names_to = "Colour",
               values_to = "Count") %>%
  mutate(Colour = gsub("_count", "", Colour)) %>%
  filter(!is.na(Count) & Count > 0)  # Remove NA and 0s

# Step 5: Define custom colours
my_colours <- c(
  Gold = "#FFD700",
  Silver = "#A9A9A9",
  White = "#D3D3D3",
  Cream = "#FFFDD0",
  Champagne = "#F7E7CE"
)

# Step 6: Plot
ggplot(family_long, aes(x = factor(family_number), y = Count, fill = Colour)) +
  geom_bar(stat = "identity") +
  labs(
    x = "Family Number",
    y = "Pearl Count",
    title = "Distribution of Pearl Colours Across Families"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5)) +
  scale_fill_manual(values = my_colours)
```



# Pearson correlation

#donor binary model
```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Gold),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#exponential curve donor binary model
```{r}
# Load required library
library(ggplot2)

# --- Step 1: Rename EBV column and ensure donor_id is a factor ---
BLUP <- BLUP %>%
  rename(EBV = solution) %>%
  mutate(donor_id = as.factor(donor_id))


# --- Step 2: Calculate observed proportion of Gold pearls per donor ---
obs_gold <- donor_dat_clean %>%
  group_by(donor_id) %>%
  summarise(
    prop_gold = mean(Gold, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Step 3: Merge EBVs with observed gold proportion ---
plot_df <- BLUP %>%
  inner_join(obs_gold, by = "donor_id") %>%
  filter(!is.na(EBV), !is.na(prop_gold))

# --- Step 4: Exponential model fit ---
plot_df <- plot_df %>%
  mutate(prop_gold_adj = pmax(prop_gold, 1e-6))  # avoid log(0)

# Starting values from log-linear regression
start_vals <- coef(lm(log(prop_gold_adj) ~ EBV, data = plot_df))
start_a <- exp(start_vals[1])
start_b <- start_vals[2]

# Fit exponential model: y = a * exp(b * x)
exp_model <- nls(prop_gold_adj ~ a * exp(b * EBV),
                 data = plot_df,
                 start = list(a = start_a, b = start_b),
                 control = list(maxiter = 200, warnOnly = TRUE))

# --- Step 5: Generate fitted curve ---
curve_df <- data.frame(EBV = seq(min(plot_df$EBV), max(plot_df$EBV), length.out = 200))
curve_df$pred <- predict(exp_model, newdata = curve_df)

# --- Step 6: Plot without point size ---
ggplot(plot_df, aes(x = EBV, y = prop_gold)) +
  geom_point(color = "darkgoldenrod", alpha = 0.8) +
  geom_line(data = curve_df, aes(x = EBV, y = pred), color = "black", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    x = "Estimated Breeding Value (EBV) for Gold",
    y = "Observed Proportion of Gold Pearls",
    title = "Donor EBV vs Observed Gold Proportion",
    subtitle = "Fitted with exponential model: y = a  exp(b  EBV)"
  ) +
  theme_minimal(base_size = 13)

# Optional: print model coefficients
cat("Fitted exponential model:\n")
print(coef(exp_model))

cor(plot_df$EBV, plot_df$prop_gold, method = "pearson", use = "complete.obs")

cor.test(plot_df$EBV, plot_df$prop_gold, method = "pearson")



```

#linear binary model
```{r}
# Calculate observed gold proportion
obs_gold <- donor_dat_clean %>%
  group_by(donor_id) %>%
  summarise(
    n = n(),
    prop_gold = mean(Gold, na.rm = TRUE)
  )

# Prepare BLUP table
BLUP$donor_id <- as.factor(BLUP$donor_id)
merged <- left_join(obs_gold, BLUP, by = "donor_id")

lm_ebv_gold <- lm(prop_gold ~ solution, data = merged)
summary(lm_ebv_gold)

library(ggplot2)

ggplot(merged, aes(x = solution, y = prop_gold)) +
  geom_point(color = "black") +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(
    x = "Estimated Breeding Value (EBV)",
    y = "Observed Proportion of Gold Pearls",
    title = "Linear Relationship between EBV and Observed Gold Proportion"
  ) +
  theme_minimal()

# Calculate Pearson correlation
cor_test <- cor.test(merged$solution, merged$prop_gold, method = "pearson")

# Print results
print(cor_test)



```




#donor quantitative model pearson correlation

```{r}

#WORKS
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- donorquant %>%
  filter(!is.na(offspring_id_donor) & !is.na(dam_donor) & !is.na(sire_donor) & !is.na(pred) & !is.na(Matched_Donor_Data_Nacre_lip_colour_Round) & !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat))

# Convert columns to appropriate types
donor_dat$offspring_id_donor <- as.factor(donor_dat$offspring_id_donor)
donor_dat$dam_donor <- as.factor(donor_dat$dam_donor)
donor_dat$sire_donor <- as.factor(donor_dat$sire_donor)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)

# Prepare pedigree
ped <- donor_dat %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modelling
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(pred, offspring_id_donor))


# Calculate average "goldness" (pred) for each donor
mean_goldness <- donor_dat %>%
  filter(!is.na(pred), !is.na(offspring_id_donor)) %>%
  group_by(offspring_id_donor) %>%
  summarise(mean_gold = mean(pred, na.rm = TRUE),
            n_pearls = n()) %>%
  ungroup()

# OPTIONAL: View summary
print(head(mean_goldness))

# Fit ASReml Gaussian animal model
mod_gold <- asreml(
  fixed = pred ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components manually
vc <- summary(mod_gold)$varcomp
print(vc)  # Print all variance components
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>% filter(donor %in% unique(donor_dat_clean$offspring_id_donor))
blup_gold$EBV <- blup_gold$solution

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)

vpredict(mod_gold,h2~V1/(V1+V2))


# Merge mean_gold with BLUPs (EBVs)
merged_df <- left_join(mean_goldness, blup_gold, by = c("offspring_id_donor" = "donor"))


# View merged table
print(merged_df)

# Merge EBV and mean_goldness
merged_df <- blup_gold %>%
  select(donor, EBV) %>%
  inner_join(mean_goldness, by = c("donor" = "offspring_id_donor"))

# Check merged result
head(merged_df)

# Load ggplot2 
library(ggplot2)

# Scatter plot + linear regression line
ggplot(merged_df, aes(x = EBV, y = mean_gold)) +
  geom_point(aes(color = "Observed Data")) +
  geom_smooth(method = "lm", se = FALSE, aes(color = "Linear Fit")) +
  scale_color_manual(values = c("Observed Data" = "black", "Linear Fit" = "blue")) +
  labs(
    x = "Estimated Breeding Value (EBV)",
    y = "Mean Predicted Colour (Goldness)",
    title = "Donor EBV vs Mean Goldness",
    color = "Legend"
  ) +
  theme_minimal()

# Pearson correlation
cor_test_result <- cor.test(merged_df$EBV, merged_df$mean_gold, method = "pearson")
print(cor_test_result)

```




# 7. Confusion Matrix (~500 pearls)



#try new good donorquant4 !!!


```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- donorquant4 %>%
  filter(!is.na(donor_offspring_id) & 
         !is.na(donor_dam) & 
         !is.na(donor_sire) & 
         !is.na(pred))

# Convert columns to appropriate types
donor_dat <- donor_dat %>%
  mutate(
    donor_offspring_id = as.factor(donor_offspring_id),
    donor_dam = as.factor(donor_dam),
    donor_sire = as.factor(donor_sire),
    Matched_Donor_Data_Nacre_lip_colour_Round = as.factor(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Flat  = as.factor(Matched_Donor_Data_Nacre_lip_colour_Flat)
  )

# Drop unused factor levels
donor_dat <- droplevels(donor_dat)

# Filter complete cases including new fixed effects
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(
    pred,
    donor_offspring_id,
    Matched_Donor_Data_Nacre_lip_colour_Round,
    Matched_Donor_Data_Nacre_lip_colour_Flat
  ))

# Add units column (required by ASReml)
donor_dat_clean$units <- 1:nrow(donor_dat_clean)

# Prepare pedigree
ped <- donor_dat_clean %>%
  distinct(donor_offspring_id, .keep_all = TRUE) %>%
  select(ID = donor_offspring_id, SIRE = donor_sire, DAM = donor_dam)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Fit ASReml Gaussian animal model with the two fixed effects
mod_gold <- asreml(
  fixed = pred ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_offspring_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components
vc <- summary(mod_gold)$varcomp
print(vc)
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Calculate heritability
vpredict(mod_gold, h2 ~ V1 / (V1 + V2))

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>%
  filter(donor %in% unique(donor_dat_clean$donor_offspring_id)) %>%
  rename(EBV = solution)

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)

```



