---
title: "everything5"
output: html_document
date: "2025-08-15"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# BINARY MODEL GOLD 
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Gold),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Gold ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Gold)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```

# BINARY MODEL SILVER
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Silver),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Silver ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Silver)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```





# BINARY MODEL WHITE
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(White),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = White ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$White)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```


# BINARY MODEL CREAM
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Cream),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Cream ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Cream)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```



# BINARY MODEL CHAMPAGNE
# TESTING THE SAME FIXED EFFECTS IN THE MODEL 

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and prepare data
host_dat <- read_csv("host_id2.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(grepl("^Gold", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Silver <- ifelse(grepl("^Silver", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$White  <- ifelse(grepl("^White", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Cream  <- ifelse(grepl("^Cream", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
host_dat$Champ  <- ifelse(grepl("^Champ", host_dat$Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)

# 3. Rename and format ID columns as factors
host_dat <- host_dat %>%
  rename(
    host_id = Shell_Metrics_GenotypeID_DART,
    Dam = host_Dam,
    Sire = host_Sire
  ) %>%
  mutate(
    host_id = as.factor(host_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 4. shell metrics and factors
host_dat$First_Op_Data_DVH_mm <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)
host_dat$First_Op_Data_Saibo_Area <- as.factor(host_dat$First_Op_Data_Saibo_Area)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_DVH_mm <- as.numeric(host_dat$Matched_Donor_Data_DVH_mm)
host_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(host_dat$Matched_Donor_Data_Shell_Weight_g)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(host_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Champ),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(host_id)
  )

# 6. Create pedigree
pedind <- host_dat_clean %>%
  select(ID = host_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

ainv <- ainverse(pedind)

# 7. Fit ASReml binary model for Gold
mod.ind <- asreml(
  fixed = Champ ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)


#vpredict(mod.ind, h2 ~ V1 / (V1 + (pi^2 / 3)))

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

vc <- summary(mod.ind)$varcomp
Va <- vc["vm(host_id, ainv)", "component"]
SE_Va <- vc["vm(host_id, ainv)", "std.error"]
Ve <- (pi^2) / 3  # Logistic residual variance

# Latent scale heritability
h2_latent <- Va / (Va + Ve)
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# Observed scale heritability
p <- mean(host_dat_clean$Champ)
z <- qnorm(p)
phi_z <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# Print results
cat("Heritability (latent scale):", round(h2_latent, 4), "\n")
cat("SE (latent scale):", round(SE_h2_latent, 4), "\n")
cat("Heritability (observed scale):", round(h2_obs, 4), "\n")
cat("SE (observed scale):", round(SE_obs, 4), "\n")

# Heritability summary
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

```



#DONOR 
GOLD BINARY MODEL # ALL FIXED EFFECTS BUT NOT GOOD

```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$First_Op_Data_DVH_mm <- as.numeric(donor_dat$First_Op_Data_DVH_mm)
donor_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(donor_dat$First_Op_Data_Nuclei_Size_mm)
donor_dat$First_Op_Data_Saibo_Area <- as.factor(donor_dat$First_Op_Data_Saibo_Area)
donor_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture <- as.factor(donor_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)
donor_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(donor_dat$Pearl_Harvest_Data_DVH_mm)
donor_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(donor_dat$Pearl_Harvest_Data_APW_mm)
donor_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(donor_dat$Pearl_Harvest_Data_Shell_Weight_g)
donor_dat$Matched_Donor_Data_DVH_mm <- as.numeric(donor_dat$Matched_Donor_Data_DVH_mm)
donor_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(donor_dat$Matched_Donor_Data_Shell_Weight_g)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Gold),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ First_Op_Data_DVH_mm + First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area
 + Pearl_Harvest_Data_Days_of_Pearl_Culture + Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g + Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```






#DONOR GOLD BINARY MODEL NO FIXED EFFECTS

```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )




# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



``` 


DONOR BINARY SILVER NO FIXED EFFECTS
```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- read_csv("donor_data2.csv")


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Silver),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Silver ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Silver, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


White no fixed effects

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- read_csv("donor_data2.csv")


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(White),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = White ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$White, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```




Cream no fixed effects

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- read_csv("donor_data2.csv")


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Cream),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Cream ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Cream, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```



Champ no fixed effects

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- read_csv("donor_data2.csv")


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Champ),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Champ ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Champ, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_champ <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


DONOR BINARY MODEL WITH FIXED EFFECTS 

#DONOR 
GOLD BINARY MODEL #with fixed effects

```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Gold),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


```{r}
# Load required library for plotting
library(ggplot2)

# --- Step 1: Prepare BLUP dataframe ---
# Rename the EBV column
BLUP <- BLUP %>%
  rename(EBV = solution) %>%
  mutate(donor_id = as.factor(donor_id))

# --- Step 2: Calculate observed proportion of Gold pearls per donor ---
obs_gold <- donor_dat_clean %>%
  group_by(donor_id) %>%
  summarise(
    n_pearls = n(),
    prop_gold = mean(Gold, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Step 3: Merge EBV and observed data ---
plot_df <- BLUP %>%
  inner_join(obs_gold, by = "donor_id") %>%
  filter(!is.na(EBV), !is.na(prop_gold))  # just in case

# --- Step 4: Exponential model fit ---
# To avoid log(0), we add a small constant only for fitting
plot_df <- plot_df %>%
  mutate(prop_gold_adj = pmax(prop_gold, 1e-6))  # use for model fitting only

# Use log-linear transformation for starting values
start_vals <- coef(lm(log(prop_gold_adj) ~ EBV, data = plot_df))
start_a <- exp(start_vals[1])
start_b <- start_vals[2]

# Fit exponential model: y = a * exp(b * x)
exp_model <- nls(prop_gold_adj ~ a * exp(b * EBV),
                 data = plot_df,
                 start = list(a = start_a, b = start_b),
                 control = list(maxiter = 200, warnOnly = TRUE))

# --- Step 5: Generate fitted curve ---
curve_df <- data.frame(EBV = seq(min(plot_df$EBV), max(plot_df$EBV), length.out = 200))
curve_df$pred <- predict(exp_model, newdata = curve_df)

# --- Step 6: Plot ---
ggplot(plot_df, aes(x = EBV, y = prop_gold)) +
  geom_point(aes(size = n_pearls), color = "darkgoldenrod", alpha = 0.8) +
  geom_line(data = curve_df, aes(x = EBV, y = pred), color = "black", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.02, 0.02))) +
  scale_size_continuous(name = "Pearls per Donor") +
  labs(
    x = "Estimated Breeding Value (EBV) for Gold",
    y = "Observed Proportion of Gold Pearls",
    title = "Donor EBV vs Observed Gold Proportion",
    subtitle = "Fitted with exponential model: y = a  exp(b  EBV)"
  ) +
  theme_minimal(base_size = 13)

# --- Optional: View model coefficients ---
cat("Fitted exponential model:\n")
print(coef(exp_model))  # a and b

```




```{r}
# Load required library
library(ggplot2)

# --- Step 1: Rename EBV column and ensure donor_id is a factor ---
BLUP <- BLUP %>%
  rename(EBV = solution) %>%
  mutate(donor_id = as.factor(donor_id))


# --- Step 2: Calculate observed proportion of Gold pearls per donor ---
obs_gold <- donor_dat_clean %>%
  group_by(donor_id) %>%
  summarise(
    prop_gold = mean(Gold, na.rm = TRUE)
  ) %>%
  ungroup()

# --- Step 3: Merge EBVs with observed gold proportion ---
plot_df <- BLUP %>%
  inner_join(obs_gold, by = "donor_id") %>%
  filter(!is.na(EBV), !is.na(prop_gold))

# --- Step 4: Exponential model fit ---
plot_df <- plot_df %>%
  mutate(prop_gold_adj = pmax(prop_gold, 1e-6))  # avoid log(0)

# Starting values from log-linear regression
start_vals <- coef(lm(log(prop_gold_adj) ~ EBV, data = plot_df))
start_a <- exp(start_vals[1])
start_b <- start_vals[2]

# Fit exponential model: y = a * exp(b * x)
exp_model <- nls(prop_gold_adj ~ a * exp(b * EBV),
                 data = plot_df,
                 start = list(a = start_a, b = start_b),
                 control = list(maxiter = 200, warnOnly = TRUE))

# --- Step 5: Generate fitted curve ---
curve_df <- data.frame(EBV = seq(min(plot_df$EBV), max(plot_df$EBV), length.out = 200))
curve_df$pred <- predict(exp_model, newdata = curve_df)

# --- Step 6: Plot without point size ---
ggplot(plot_df, aes(x = EBV, y = prop_gold)) +
  geom_point(color = "darkgoldenrod", alpha = 0.8) +
  geom_line(data = curve_df, aes(x = EBV, y = pred), color = "black", linewidth = 1) +
  scale_y_continuous(limits = c(0, 1), expand = expansion(mult = c(0.02, 0.02))) +
  labs(
    x = "Estimated Breeding Value (EBV) for Gold",
    y = "Observed Proportion of Gold Pearls",
    title = "Donor EBV vs Observed Gold Proportion",
    subtitle = "Fitted with exponential model: y = a  exp(b  EBV)"
  ) +
  theme_minimal(base_size = 13)

# Optional: print model coefficients
cat("Fitted exponential model:\n")
print(coef(exp_model))

cor(plot_df$EBV, plot_df$prop_gold, method = "pearson", use = "complete.obs")

cor.test(plot_df$EBV, plot_df$prop_gold, method = "pearson")



```



#DONOR SILVER BINARY MODEL # fixed effects but not good 

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Silver),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Silver ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Silver, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```


#DONOR 
WHITE BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)


# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(White),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = White ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$White, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```






#DONOR 
CREAM BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)
donor_dat$Matched_Donor_Data_DVH_mm <- as.numeric(donor_dat$Matched_Donor_Data_DVH_mm)
donor_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(donor_dat$Matched_Donor_Data_Shell_Weight_g)



# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Cream),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Cream ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat + Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Cream, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```




#DONOR 
CHAMP BINARY MODEL # FIXED EFFECTS

```{r}
# DONOR 

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
#donor_dat <- read_csv("donor_data.csv")
donor_dat <- donor_data2


# 3. Rename and format ID columns as factors
donor_dat <- donor_dat %>%
  rename(
    donor_id = Matched_Donor_Data_Saibo_No,
    Dam = `Candidate mother ID`,
    Sire = `Candidate father ID`
  ) %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam     = as.factor(Dam),
    Sire    = as.factor(Sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
     )

# 4. shell metrics and factors
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat <- as.factor(donor_dat$Matched_Donor_Data_Nacre_lip_colour_Flat)




# 5. Remove rows with missing values
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Champ),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id)
  )


# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Champ ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# Calculating heritability
#vpredict(mod.ind,h2~V1/(V1+V2))

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- (pi^2)/3
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Champ, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
#actual_ids <- unique(donor_dat_clean$donor_id)
#BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)



```






use this vpredict host

```{r}

# HOST QUANTITATIVE  CLEAN VERSION (NO DISTINCT, CONVERGED TWICE, NO HERITABILITY SECTION)

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- hostquant %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# Convert columns to appropriate types
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# Prepare pedigree (no distinct used)
ped <- host_dat %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)

# Update twice to ensure convergence and standard errors
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)

# Extract and print variance components
vc <- summary(mod_host)$varcomp
print(vc)

# Calculating heritability
vpredict(mod_host,h2~V1/(V1+V2))



# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>%
  filter(host %in% unique(host_dat_clean$offspring_id_host)) %>%
  mutate(EBV = solution)

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```




# DONOR QUANTITATIVE GOOD

```{r}
#WORKS
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- donorquant %>%
  filter(!is.na(offspring_id_donor) & !is.na(dam_donor) & !is.na(sire_donor) & !is.na(pred))

# Convert columns to appropriate types
donor_dat$offspring_id_donor <- as.factor(donor_dat$offspring_id_donor)
donor_dat$dam_donor <- as.factor(donor_dat$dam_donor)
donor_dat$sire_donor <- as.factor(donor_dat$sire_donor)

# Prepare pedigree
ped <- donor_dat %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modelling
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(pred, offspring_id_donor))

# Fit ASReml Gaussian animal model
mod_gold <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components manually
vc <- summary(mod_gold)$varcomp
print(vc)  # Print all variance components
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>% filter(donor %in% unique(donor_dat_clean$offspring_id_donor))
blup_gold$EBV <- blup_gold$solution

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)

vpredict(mod_gold,h2~V1/(V1+V2))

```


```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- donorquant %>%
  filter(!is.na(offspring_id_donor) & 
         !is.na(dam_donor) & 
         !is.na(sire_donor) & 
         !is.na(pred))

# Convert columns to appropriate types
donor_dat <- donor_dat %>%
  mutate(
    offspring_id_donor = as.factor(offspring_id_donor),
    dam_donor = as.factor(dam_donor),
    sire_donor = as.factor(sire_donor),
    Matched_Donor_Data_Nacre_lip_colour_Round = as.factor(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Flat  = as.factor(Matched_Donor_Data_Nacre_lip_colour_Flat)
  )

# Drop unused factor levels
donor_dat <- droplevels(donor_dat)

# Filter complete cases including new fixed effects
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(
    pred,
    offspring_id_donor,
    Matched_Donor_Data_Nacre_lip_colour_Round,
    Matched_Donor_Data_Nacre_lip_colour_Flat
  ))

# Add units column (required by ASReml)
donor_dat_clean$units <- 1:nrow(donor_dat_clean)

# Prepare pedigree
ped <- donor_dat_clean %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Fit ASReml Gaussian animal model with the two fixed effects
mod_gold <- asreml(
  fixed = pred ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components
vc <- summary(mod_gold)$varcomp
print(vc)
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Calculate heritability
vpredict(mod_gold, h2 ~ V1 / (V1 + V2))

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>%
  filter(donor %in% unique(donor_dat_clean$offspring_id_donor)) %>%
  rename(EBV = solution)

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)

```

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- hostquant %>%
  filter(!is.na(offspring_id_host) & 
         !is.na(dam_host) & 
         !is.na(sire_host) & 
         !is.na(pred))

# Convert ID columns to factor
host_dat <- host_dat %>%
  mutate(
    offspring_id_host = as.factor(offspring_id_host),
    dam_host = as.factor(dam_host),
    sire_host = as.factor(sire_host),
    
    # Fixed effects
    First_Op_Data_DVH_mm = as.numeric(First_Op_Data_DVH_mm),
    First_Op_Data_Nuclei_Size_mm = as.numeric(First_Op_Data_Nuclei_Size_mm),
    First_Op_Data_Saibo_Area = as.factor(First_Op_Data_Saibo_Area),
    Pearl_Harvest_Data_Days_of_Pearl_Culture = as.factor(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    Pearl_Harvest_Data_DVH_mm = as.numeric(Pearl_Harvest_Data_DVH_mm),
    Pearl_Harvest_Data_APW_mm = as.numeric(Pearl_Harvest_Data_APW_mm),
    Pearl_Harvest_Data_Shell_Weight_g = as.numeric(Pearl_Harvest_Data_Shell_Weight_g),
    Matched_Donor_Data_DVH_mm = as.numeric(Matched_Donor_Data_DVH_mm),
    Matched_Donor_Data_Shell_Weight_g = as.numeric(Matched_Donor_Data_Shell_Weight_g),
    Matched_Donor_Data_Nacre_lip_colour_Round = as.factor(Matched_Donor_Data_Nacre_lip_colour_Round),
    Matched_Donor_Data_Nacre_lip_colour_Flat  = as.factor(Matched_Donor_Data_Nacre_lip_colour_Flat)
  )

# Drop unused factor levels
host_dat <- droplevels(host_dat)

# Filter complete cases for all model variables
host_dat_clean <- host_dat %>%
  filter(complete.cases(
    pred,
    offspring_id_host,
    First_Op_Data_DVH_mm,
    First_Op_Data_Nuclei_Size_mm,
    First_Op_Data_Saibo_Area,
    Pearl_Harvest_Data_Days_of_Pearl_Culture,
    Pearl_Harvest_Data_DVH_mm,
    Pearl_Harvest_Data_APW_mm,
    Pearl_Harvest_Data_Shell_Weight_g,
    Matched_Donor_Data_DVH_mm,
    Matched_Donor_Data_Shell_Weight_g,
    Matched_Donor_Data_Nacre_lip_colour_Round,
    Matched_Donor_Data_Nacre_lip_colour_Flat
  ))

# Add 'units' column
host_dat_clean$units <- 1:nrow(host_dat_clean)

# Prepare pedigree (no distinct used)
ped <- host_dat_clean %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Fit ASReml Gaussian animal model with fixed effects
mod_host <- asreml(
  fixed = pred ~ First_Op_Data_DVH_mm +
                 First_Op_Data_Nuclei_Size_mm +
                 First_Op_Data_Saibo_Area +
                 Pearl_Harvest_Data_Days_of_Pearl_Culture +
                 Pearl_Harvest_Data_DVH_mm +
                 Pearl_Harvest_Data_APW_mm +
                 Pearl_Harvest_Data_Shell_Weight_g +
                 Matched_Donor_Data_DVH_mm +
                 Matched_Donor_Data_Shell_Weight_g +
                 Matched_Donor_Data_Nacre_lip_colour_Round +
                 Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)

# Update model to ensure convergence
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)
mod_host <- update(mod_host)

# Extract and print variance components
vc <- summary(mod_host)$varcomp
print(vc)

# Heritability (latent, via vpredict)
vpredict(mod_host, h2 ~ V1 / (V1 + V2))

# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>%
  filter(host %in% unique(host_dat_clean$offspring_id_host)) %>%
  mutate(EBV = solution)

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```

