---
title: "se2"
output: html_document
date: "2025-07-17"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## TRADITIONAL HERITABILITY ESTIMATES AND BLUP (EBV)

#HOST GOLD

```{r}


# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
host_dat <- read_csv("host_id.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
host_dat$Silver <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
host_dat$White  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
host_dat$Cream  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
host_dat$Champ  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 3. Format pedigree columns as factors
host_dat$host_id <- as.factor(host_dat$host_id)
host_dat$Dam     <- as.factor(host_dat$Dam)
host_dat$Sire    <- as.factor(host_dat$Sire)

# 4. Ensure shell metrics are numeric
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_APW_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_APW_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)

# 5. Remove rows with missing values in relevant columns
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Gold),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(Pearl_Harvest_Data_APW_mm),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(host_id)
  )

# 6. Create pedigree and inverse relationship matrix
pedind <- host_dat_clean[, c("host_id", "Sire", "Dam")]
colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))

ainv <- ainverse(pedind)

# 7. Fit animal model for Gold (binary trait)
mod.ind <- asreml(
  fixed = Gold ~ Pearl_Harvest_Data_DVH_mm + Pearl_Harvest_Data_APW_mm + Pearl_Harvest_Data_Shell_Weight_g,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)



# Latent scale tells you the heritability on a theoretical continuous trait that gives rise to the binary outcome.

# Observed scale tells you how much of the 0/1 variation is due to genetics — and is usually much lower, especially when the trait is rare

# --- Extract variance components ---
vc     <- summary(mod.ind)$varcomp
Va     <- vc["vm(host_id, ainv)", "component"]     # Additive genetic variance
SE_Va  <- vc["vm(host_id, ainv)", "std.error"]     # Standard error of Va
Ve     <- 1                                        # Residual variance for logistic models

# --- Latent-scale heritability ---
h2_latent     <- Va / (Va + Ve)
dVa           <- Ve / (Va + Ve)^2                 # Derivative for delta method
SE_h2_latent  <- SE_Va * dVa

cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")

# --- Observed-scale heritability ---
p      <- mean(host_dat_clean$Gold, na.rm = TRUE)  # Trait prevalence
z      <- qnorm(p)
phi_z  <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs    <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_obs, "\n")

# --- Summary table of heritability and standard errors ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)


# Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

#BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

#BLUP$prob_gold


```

# HOST SILVER

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
host_dat <- read_csv("host_id.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
host_dat$Silver <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
host_dat$White  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
host_dat$Cream  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
host_dat$Champ  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 3. Format pedigree columns as factors
host_dat$host_id <- as.factor(host_dat$host_id)
host_dat$Dam     <- as.factor(host_dat$Dam)
host_dat$Sire    <- as.factor(host_dat$Sire)
host_dat$First_Op_Data_Saibo_Area    <- as.factor(host_dat$First_Op_Data_Saibo_Area)

# 4. Ensure shell metrics are numeric
host_dat$First_Op_Data_Nuclei_Size_mm         <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)


# 5. Remove rows with missing values in relevant columns
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Silver),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(host_id)
  )

# 6. Create pedigree and inverse relationship matrix
pedind <- host_dat_clean[, c("host_id", "Sire", "Dam")]
colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))

ainv <- ainverse(pedind)

# 7. Fit animal model for Silver (binary trait)
mod.ind <- asreml(
  fixed = Silver ~ First_Op_Data_Nuclei_Size_mm + First_Op_Data_Saibo_Area,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)



# calculate heritability on the observed scale from your animal model for a binary trait. This gives a more interpretable estimate in terms of the actual 0/1 outcomes (i.e., real-world expression of color).


# Latent scale tells you the heritability on a theoretical continuous trait that gives rise to the binary outcome.

# Observed scale tells you how much of the 0/1 variation is due to genetics — and is usually much lower, especially when the trait is rare

# --- Extract variance components ---
vc     <- summary(mod.ind)$varcomp
Va     <- vc["vm(host_id, ainv)", "component"]     # Additive genetic variance
SE_Va  <- vc["vm(host_id, ainv)", "std.error"]     # Standard error of Va
Ve     <- 1                                        # Residual variance for logistic models

# --- Latent-scale heritability ---
h2_latent     <- Va / (Va + Ve)
dVa           <- Ve / (Va + Ve)^2                 # Derivative for delta method
SE_h2_latent  <- SE_Va * dVa

cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")

# --- Observed-scale heritability ---
p      <- mean(host_dat_clean$Silver, na.rm = TRUE)  # Trait prevalence
z      <- qnorm(p)
phi_z  <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs    <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_obs, "\n")

# --- Summary table of heritability and standard errors ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# 11. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

#BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

#BLUP$prob_silver

```

# HOST WHITE

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
host_dat <- read_csv("host_id.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
host_dat$Silver <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
host_dat$White  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
host_dat$Cream  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
host_dat$Champ  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 3. Format pedigree columns as factors
host_dat$host_id <- as.factor(host_dat$host_id)
host_dat$Dam     <- as.factor(host_dat$Dam)
host_dat$Sire    <- as.factor(host_dat$Sire)
host_dat$First_Op_Data_Saibo_Area    <- as.factor(host_dat$First_Op_Data_Saibo_Area)

# 4. Ensure shell metrics are numeric
host_dat$Pearl_Harvest_Data_DVH_mm         <- as.numeric(host_dat$Pearl_Harvest_Data_DVH_mm)
host_dat$Pearl_Harvest_Data_Shell_Weight_g         <- as.numeric(host_dat$Pearl_Harvest_Data_Shell_Weight_g)

# 5. Remove rows with missing values in relevant columns
host_dat_clean <- host_dat %>%
  filter(
    !is.na(White),
    !is.na(Pearl_Harvest_Data_DVH_mm),
    !is.na(First_Op_Data_Saibo_Area),
    !is.na(Pearl_Harvest_Data_Shell_Weight_g),
    !is.na(host_id)
  )

# 6. Create pedigree and inverse relationship matrix
pedind <- host_dat_clean[, c("host_id", "Sire", "Dam")]
colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))

ainv <- ainverse(pedind)

# 7. Fit animal model for White (binary trait)
mod.ind <- asreml(
  fixed = White ~ Pearl_Harvest_Data_DVH_mm + First_Op_Data_Saibo_Area + Pearl_Harvest_Data_Shell_Weight_g,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)



# calculate heritability on the observed scale from your animal model for a binary trait. This gives a more interpretable estimate in terms of the actual 0/1 outcomes (i.e., real-world expression of color).


# Latent scale tells you the heritability on a theoretical continuous trait that gives rise to the binary outcome.

# Observed scale tells you how much of the 0/1 variation is due to genetics — and is usually much lower, especially when the trait is rare

# --- Extract variance components ---
vc     <- summary(mod.ind)$varcomp
Va     <- vc["vm(host_id, ainv)", "component"]     # Additive genetic variance
SE_Va  <- vc["vm(host_id, ainv)", "std.error"]     # Standard error of Va
Ve     <- 1                                        # Residual variance for logistic models

# --- Latent-scale heritability ---
h2_latent     <- Va / (Va + Ve)
dVa           <- Ve / (Va + Ve)^2                 # Derivative for delta method
SE_h2_latent  <- SE_Va * dVa

cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")

# --- Observed-scale heritability ---
p      <- mean(host_dat_clean$White, na.rm = TRUE)  # Trait prevalence
z      <- qnorm(p)
phi_z  <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs    <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_obs, "\n")

# --- Summary table of heritability and standard errors ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)


# 11. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

#BLUP$prob_white <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

#BLUP$prob_white

```

# HOST CREAM

```{r}

#Scaling does not affect BLUPs or heritability structure
#Random effects and ainv matrix are independent of the scale of the fixed effects.
#So your heritability estimates and BLUPs remain interpretable — you just get a more numerically stable model.
# I noticed when I do the fixed effects separately in the model the BLUP works, the First_Op_Data_Nuclei_Size_mm values are way smaller and have decimal pointsthan the First_Op_Data_DVH_mm
# also the First_Op_Data_DVH_mm has some outliers 160, 90 but the average is around 100-150, even when i removed the BLUP is still very high, i tried quantile too but it doesn't work, thus scale is best

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
host_dat <- read_csv("host_id.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
host_dat$Silver <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
host_dat$White  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
host_dat$Cream  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
host_dat$Champ  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 3. Format pedigree columns as factors
host_dat$host_id <- as.factor(host_dat$host_id)
host_dat$Dam     <- as.factor(host_dat$Dam)
host_dat$Sire    <- as.factor(host_dat$Sire)

# 4. Ensure predictors are numeric
host_dat$First_Op_Data_DVH_mm         <- as.numeric(host_dat$First_Op_Data_DVH_mm)
host_dat$First_Op_Data_Nuclei_Size_mm <- as.numeric(host_dat$First_Op_Data_Nuclei_Size_mm)

# 5. Remove rows with missing values
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Cream),
    !is.na(First_Op_Data_DVH_mm),
    !is.na(First_Op_Data_Nuclei_Size_mm),
    !is.na(host_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 6. Create pedigree and inverse relationship matrix
pedind <- host_dat_clean[, c("host_id", "Sire", "Dam")]
colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))
ainv <- ainverse(pedind)

# 7. Scale predictors
host_dat_clean <- host_dat_clean %>%
  mutate(
    DVH_scaled     = scale(First_Op_Data_DVH_mm),
    Nuclei_scaled  = scale(First_Op_Data_Nuclei_Size_mm)
  )

# 8. Fit animal model for Cream (binary trait)
mod.ind <- asreml(
  fixed    = Cream ~ DVH_scaled + Nuclei_scaled,
  random   = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data     = host_dat_clean,
  family   = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)



# Latent scale tells you the heritability on a theoretical continuous trait that gives rise to the binary outcome.

# Observed scale tells you how much of the 0/1 variation is due to genetics — and is usually much lower, especially when the trait is rare

# --- Extract variance components ---
vc     <- summary(mod.ind)$varcomp
Va     <- vc["vm(host_id, ainv)", "component"]     # Additive genetic variance
SE_Va  <- vc["vm(host_id, ainv)", "std.error"]     # Standard error of Va
Ve     <- 1                                        # Residual variance for logistic models

# --- Latent-scale heritability ---
h2_latent     <- Va / (Va + Ve)
dVa           <- Ve / (Va + Ve)^2                 # Derivative for delta method
SE_h2_latent  <- SE_Va * dVa

cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")

# --- Observed-scale heritability ---
p      <- mean(host_dat_clean$Cream, na.rm = TRUE)  # Trait prevalence
z      <- qnorm(p)
phi_z  <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs    <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_obs, "\n")

# --- Summary table of heritability and standard errors ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)

# 11. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$host_id <- rownames(BLUP)
head(BLUP)

```

# HOST cHAMPAGNE

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
host_dat <- read_csv("host_id.csv")

# 2. Create binary colour traits
host_dat$Gold   <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Gold"), 1, 0)
host_dat$Silver <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Silver"), 1, 0)
host_dat$White  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "White"), 1, 0)
host_dat$Cream  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Cream"), 1, 0)
host_dat$Champ  <- ifelse(startsWith(host_dat$Pearl_Grading_Data_Colour, "Champ"), 1, 0)

# 3. Format pedigree columns as factors
host_dat$host_id <- as.factor(host_dat$host_id)
host_dat$Dam     <- as.factor(host_dat$Dam)
host_dat$Sire    <- as.factor(host_dat$Sire)
host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture    <- as.factor(host_dat$Pearl_Harvest_Data_Days_of_Pearl_Culture)




# 5. Remove rows with missing values in relevant columns
host_dat_clean <- host_dat %>%
  filter(
    !is.na(Champ),
    !is.na(Pearl_Harvest_Data_Days_of_Pearl_Culture),
    !is.na(host_id)
  )

# 6. Create pedigree and inverse relationship matrix
pedind <- host_dat_clean[, c("host_id", "Sire", "Dam")]
colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))

ainv <- ainverse(pedind)

# 7. Fit animal model for champ (binary trait)
mod.ind <- asreml(
  fixed = Champ ~ Pearl_Harvest_Data_Days_of_Pearl_Culture,
  random = ~ vm(host_id, ainv),
  residual = ~ idv(units),
  data = host_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)



# calculate heritability on the observed scale from your animal model for a binary trait. This gives a more interpretable estimate in terms of the actual 0/1 outcomes (i.e., real-world expression of color).

# Latent scale tells you the heritability on a theoretical continuous trait that gives rise to the binary outcome.

# Observed scale tells you how much of the 0/1 variation is due to genetics — and is usually much lower, especially when the trait is rare

# --- Extract variance components ---
vc     <- summary(mod.ind)$varcomp
Va     <- vc["vm(host_id, ainv)", "component"]     # Additive genetic variance
SE_Va  <- vc["vm(host_id, ainv)", "std.error"]     # Standard error of Va
Ve     <- 1                                        # Residual variance for logistic models

# --- Latent-scale heritability ---
h2_latent     <- Va / (Va + Ve)
dVa           <- Ve / (Va + Ve)^2                 # Derivative for delta method
SE_h2_latent  <- SE_Va * dVa

cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")

# --- Observed-scale heritability ---
p      <- mean(host_dat_clean$Champ, na.rm = TRUE)  # Trait prevalence
z      <- qnorm(p)
phi_z  <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_obs    <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_obs, "\n")

# --- Summary table of heritability and standard errors ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_obs)
)

print(herit_summary)


# 11. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
head(BLUP)

#BLUP$prob_champ <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

#BLUP$prob_champ

```

## DONOR

#DONOR GOLD

```{r}
# DONOR GOLD

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
donor_dat <- read_csv("donor_data.csv")

# 2. Fix column names and convert to factors
donor_dat <- donor_dat %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam = as.factor(dam),
    Sire = as.factor(sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# 4. Filter for complete cases required for modelling
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Gold),
    !is.na(donor_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Fit the animal model
mod.ind <- asreml(
  fixed = Gold ~ 1,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)


# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- 1
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Gold, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs    <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

# Print the table
print(herit_summary)


# 13. Extract BLUPs (genetic merit)
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))

# Filter only donor IDs used in model
actual_ids <- unique(donor_dat_clean$donor_id)
BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
BLUP$prob_gold <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)

```

# DONOR SILVER

```{r}
# DONOR SILVER

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
donor_dat <- read_csv("donor_data.csv")

# 2. Fix column names and convert to factors
donor_dat <- donor_dat %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam = as.factor(dam),
    Sire = as.factor(sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (start with string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# 4. Filter for complete cases required for modelling
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Silver),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(donor_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 5. Create a pedigree with unique donor IDs only
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam)

colnames(pedind) <- c("ID", "SIRE", "DAM")
pedind <- data.frame(lapply(pedind, as.factor))  # Ensure factors

# 6. Generate the inverse relationship matrix (A-inverse)
ainv <- ainverse(pedind)

# 7. Ensure fixed effect is a factor
donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(
  donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round
)

# 8. Fit the animal model
mod.ind <- asreml(
  fixed = Silver ~ Matched_Donor_Data_Nacre_lip_colour_Round,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- 1
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p <- mean(donor_dat_clean$Silver, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# Create a summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)

print(herit_summary)

# 13. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))
actual_ids <- unique(donor_dat_clean$donor_id)
BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# Optional: calculate probabilities
BLUP$prob_silver <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)

```

# DONOR WHITE

```{r}
# DONOR WHITE

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
donor_dat <- read_csv("donor_data.csv")

# 2. Convert relevant columns to factors
donor_dat <- donor_dat %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam = as.factor(dam),
    Sire = as.factor(sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour (starting string match)
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ",  Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# 4. Filter for complete cases required for modelling
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(White),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(donor_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 5. Create pedigree
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam) %>%
  rename(ID = donor_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

# 6. Generate A-inverse matrix
ainv <- ainverse(pedind)

# 7. Ensure fixed effect is factor
donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(
  donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round
)

# 8. Fit animal model
mod.ind <- asreml(
  fixed = White ~ Matched_Donor_Data_Nacre_lip_colour_Round,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- 1
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa          <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using trait prevalence
p      <- mean(donor_dat_clean$White, na.rm = TRUE)
z      <- qnorm(p)
phi_z  <- dnorm(z)
h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# --- Summary table ---
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)
print(herit_summary)

# 13. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))
actual_ids <- as.character(unique(donor_dat_clean$donor_id))
BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# 14. Calculate probabilities
BLUP$prob_white <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)

```

# DONOR CREAM

```{r}
# DONOR CREAM

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
donor_dat <- read_csv("donor_data.csv")

# 2. Convert relevant columns to factors
donor_dat <- donor_dat %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam = as.factor(dam),
    Sire = as.factor(sire)
  )

# 3. Ensure shell metrics are numeric
donor_dat$Matched_Donor_Data_DVH_mm <- as.numeric(donor_dat$Matched_Donor_Data_DVH_mm)
donor_dat$Matched_Donor_Data_Shell_Weight_g <- as.numeric(donor_dat$Matched_Donor_Data_Shell_Weight_g)

# 4. Create binary traits from Pearl_Grading_Data_Colour
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# 5. Filter for complete cases
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Cream),
    !is.na(Matched_Donor_Data_DVH_mm),
    !is.na(Matched_Donor_Data_Shell_Weight_g),
    !is.na(donor_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 6. Create pedigree
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam) %>%
  rename(ID = donor_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

# 7. Generate A-inverse matrix
ainv <- ainverse(pedind)

# 8. Fit animal model
mod.ind <- asreml(
  fixed = Cream ~ Matched_Donor_Data_DVH_mm + Matched_Donor_Data_Shell_Weight_g,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- 1
h2_latent <- Va / (Va + Ve)

# SE using delta method
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using Cream prevalence
p <- mean(donor_dat_clean$Cream, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# 13. Summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)
print(herit_summary)

# 14. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))
actual_ids <- as.character(unique(donor_dat_clean$donor_id))
BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# 15. Calculate predicted probabilities (optional)
BLUP$prob_cream <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)

```

# DONOR CHAMPAGNE

```{r}
# DONOR CHAMP

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load data
donor_dat <- read_csv("donor_data.csv")

# 2. Convert relevant columns to factors
donor_dat <- donor_dat %>%
  mutate(
    donor_id = as.factor(donor_id),
    Dam = as.factor(dam),
    Sire = as.factor(sire)
  )

# 3. Create binary traits from Pearl_Grading_Data_Colour
donor_dat <- donor_dat %>%
  mutate(
    Gold   = ifelse(grepl("^Gold", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Silver = ifelse(grepl("^Silver", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    White  = ifelse(grepl("^White", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Cream  = ifelse(grepl("^Cream", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0),
    Champ  = ifelse(grepl("^Champ", Pearl_Grading_Data_Colour, ignore.case = TRUE), 1, 0)
  )

# 4. Filter for complete cases required for modelling
donor_dat_clean <- donor_dat %>%
  filter(
    !is.na(Champ),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Round),
    !is.na(Matched_Donor_Data_Nacre_lip_colour_Flat),
    !is.na(donor_id),
    !is.na(Sire),
    !is.na(Dam)
  )

# 5. Create a pedigree
pedind <- donor_dat_clean %>%
  distinct(donor_id, .keep_all = TRUE) %>%
  select(donor_id, Sire, Dam) %>%
  rename(ID = donor_id, SIRE = Sire, DAM = Dam) %>%
  mutate(across(everything(), as.factor))

# 6. Generate inverse relationship matrix
ainv <- ainverse(pedind)

# 7. Ensure fixed effects are factors
donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round <- as.factor(donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Round)
donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Flat  <- as.factor(donor_dat_clean$Matched_Donor_Data_Nacre_lip_colour_Flat)

# 8. Fit the animal model
mod.ind <- asreml(
  fixed = Champ ~ Matched_Donor_Data_Nacre_lip_colour_Round + Matched_Donor_Data_Nacre_lip_colour_Flat,
  random = ~ vm(donor_id, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean,
  family = asr_binomial(link = "logit")
)

# 8. Update and extract variance components
mod.ind <- update.asreml(mod.ind)
summary(mod.ind)$varcomp

# 8. Update model (again) to ensure convergence
mod.ind <- update.asreml(mod.ind)

# 9. Extract variance components manually
vc <- summary(mod.ind)$varcomp
print(vc)

# 10. Calculate heritability (latent scale)
Va    <- vc["vm(donor_id, ainv)", "component"]
SE_Va <- vc["vm(donor_id, ainv)", "std.error"]
Ve    <- 1
h2_latent <- Va / (Va + Ve)

# Standard error using delta method
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 11. Convert to observed scale using Champ prevalence
p <- mean(donor_dat_clean$Champ, na.rm = TRUE)
z <- qnorm(p)
phi_z <- dnorm(z)

h2_obs <- (h2_latent * phi_z^2) / (p * (1 - p))
SE_h2_obs <- (SE_h2_latent * phi_z^2) / (p * (1 - p))

# 12. Print heritability estimates and SEs
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# 13. Summary table
herit_summary <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)
print(herit_summary)

# 14. Extract BLUPs
BLUP <- as.data.frame(summary(mod.ind, coef = TRUE)$coef.random)
BLUP$donor_id <- sub(".*donor_id, ainv\\)_", "", rownames(BLUP))
actual_ids <- as.character(unique(donor_dat_clean$donor_id))
BLUP <- BLUP %>% filter(donor_id %in% actual_ids)

# 15. Calculate predicted probabilities (optional)
BLUP$prob_champ <- exp(BLUP$solution) / (1 + exp(BLUP$solution))

head(BLUP)

```

## QUANTITATIVE HERITABILITY ESTIMATE AND BLUP (EBV)

# HOST QUANTITATIVE

```{r}
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and clean host data
host_dat <- read_csv("hostquant.csv") %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# 2. Convert ID columns to factors
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# 3. Prepare pedigree and create inverse relationship matrix
ped_host <- host_dat %>%
  distinct(offspring_id_host, .keep_all = TRUE) %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped_host <- data.frame(lapply(ped_host, as.factor))
ainv_host <- ainverse(ped_host)

# 4. Filter complete cases for modeling
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# 5. Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_host, ainv_host),
  residual = ~ idv(units),
  data = host_dat_clean
)
mod_host <- update(mod_host)

# 6. Extract variance components
vc_host <- summary(mod_host)$varcomp
print(vc_host)

# 7. Calculate heritability (latent scale)
Va_host <- vc_host["vm(offspring_id_host, ainv_host)", "component"]
Ve_host <- vc_host["units!R", "component"]
h2_latent_host <- Va_host / (Va_host + Ve_host)

# 8. Calculate heritability (observed scale)
pred_var_host <- var(host_dat_clean$pred, na.rm = TRUE)
h2_observed_host <- Va_host / pred_var_host

# 9. Print heritability estimates
cat("Gold heritability (latent scale):", round(h2_latent_host, 4), "\n")
cat("Gold heritability (observed scale):", round(h2_observed_host, 4), "\n")

# 10. Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>% filter(host %in% unique(host_dat_clean$offspring_id_host))
blup_host$EBV <- blup_host$solution

# 11. View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```

```{r}
# HOST QUANTITATIVE HERITABILITY + SE

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# 1. Load and clean host data
host_dat <- read_csv("hostquant.csv") %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# 2. Convert ID columns to factors
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# 3. Prepare pedigree and inverse relationship matrix
ped_host <- host_dat %>%
  distinct(offspring_id_host, .keep_all = TRUE) %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped_host <- data.frame(lapply(ped_host, as.factor))
ainv_host <- ainverse(ped_host)

# 4. Filter complete cases
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# 5. Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed    = pred ~ 1,
  random   = ~ vm(offspring_id_host, ainv_host),
  residual = ~ idv(units),
  data     = host_dat_clean
)

# 6. Ensure convergence
mod_host <- update(mod_host)
mod_host <- update(mod_host)

# 7. Extract variance components
vc_host <- summary(mod_host)$varcomp
print(vc_host)

# 8. Heritability (latent scale)
Va    <- vc_host["vm(offspring_id_host, ainv_host)", "component"]
SE_Va <- vc_host["vm(offspring_id_host, ainv_host)", "std.error"]
Ve    <- vc_host["units!R", "component"]

h2_latent <- Va / (Va + Ve)

# Delta method for standard error
dVa <- Ve / (Va + Ve)^2
SE_h2_latent <- SE_Va * dVa

# 9. Heritability (observed scale)
Vp_obs <- var(host_dat_clean$pred, na.rm = TRUE)
h2_obs <- Va / Vp_obs
SE_h2_obs <- SE_Va / Vp_obs  # Approximate SE on observed scale

# 10. Output
cat("Heritability (latent scale):       ", h2_latent, "\n")
cat("Standard Error (latent scale):     ", SE_h2_latent, "\n")
cat("Heritability (observed scale):     ", h2_obs, "\n")
cat("Standard Error (observed scale):   ", SE_h2_obs, "\n")

# 11. Summary table
herit_table <- data.frame(
  Scale = c("Latent", "Observed"),
  Heritability = c(h2_latent, h2_obs),
  Std_Error = c(SE_h2_latent, SE_h2_obs)
)
print(herit_table)

```

# DONOR QUANTITATIVE issue

```{r}

#issues
# DONOR QUANTITATIVE (Gaussian model, no latent/observed scale conversion)

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- read_csv("donorquant.csv") %>%
  filter(!is.na(offspring_id_donor) & !is.na(dam_donor) & !is.na(sire_donor) & !is.na(pred))

# Convert columns to appropriate types
donor_dat$offspring_id_donor <- as.factor(donor_dat$offspring_id_donor)
donor_dat$dam_donor <- as.factor(donor_dat$dam_donor)
donor_dat$sire_donor <- as.factor(donor_dat$sire_donor)

# Prepare pedigree and inverse relationship matrix
ped <- donor_dat %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor) %>%
  mutate(across(everything(), as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(pred, offspring_id_donor))

# Fit ASReml Gaussian animal model
mod_gold <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components
vc <- summary(mod_gold)$varcomp
print(vc)

# Manually extract components
Va    <- vc["vm(offspring_id_donor, ainv)", "component"]
SE_Va <- vc["vm(offspring_id_donor, ainv)", "std.error"]
Ve    <- vc["units!R", "component"]

# Total phenotypic variance
Vp <- Va + Ve

# Heritability
h2 <- Va / Vp

# Standard error using delta method
dVa <- (Vp - Va) / (Vp^2)
SE_h2 <- SE_Va * dVa

# Print results
cat("Heritability (Gaussian model):     ", h2, "\n")
cat("Standard Error (heritability):     ", SE_h2, "\n")

# Optional summary table
herit_summary <- data.frame(
  Heritability = h2,
  Std_Error = SE_h2
)
print(herit_summary)

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>% filter(donor %in% unique(donor_dat_clean$offspring_id_donor))
blup_gold$EBV <- blup_gold$solution

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)


#The first block uses this line for standard error:


#dVa <- (Vp - Va) / (Vp^2)
#SE_h2 <- SE_Va * dVa
#Which ignores the uncertainty in the residual variance (Ve), making it less accurate.

```

# DONOR QUANTITATIVE

```{r}
#WORKS
# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
donor_dat <- read_csv("donorquant.csv") %>%
  filter(!is.na(offspring_id_donor) & !is.na(dam_donor) & !is.na(sire_donor) & !is.na(pred))

# Convert columns to appropriate types
donor_dat$offspring_id_donor <- as.factor(donor_dat$offspring_id_donor)
donor_dat$dam_donor <- as.factor(donor_dat$dam_donor)
donor_dat$sire_donor <- as.factor(donor_dat$sire_donor)

# Prepare pedigree
ped <- donor_dat %>%
  distinct(offspring_id_donor, .keep_all = TRUE) %>%
  select(ID = offspring_id_donor, SIRE = sire_donor, DAM = dam_donor)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
donor_dat_clean <- donor_dat %>%
  filter(complete.cases(pred, offspring_id_donor))

# Fit ASReml Gaussian animal model
mod_gold <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_donor, ainv),
  residual = ~ idv(units),
  data = donor_dat_clean
)
mod_gold <- update(mod_gold)

# Extract variance components manually
vc <- summary(mod_gold)$varcomp
print(vc)  # Print all variance components
Va <- vc["vm(offspring_id_donor, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Extract BLUPs
blup_gold <- as.data.frame(summary(mod_gold, coef = TRUE)$coef.random)
blup_gold$donor <- gsub("^.*\\)_", "", rownames(blup_gold))
blup_gold <- blup_gold %>% filter(donor %in% unique(donor_dat_clean$offspring_id_donor))
blup_gold$EBV <- blup_gold$solution

# View top BLUPs
head(blup_gold[order(-blup_gold$EBV), ], 10)



```

# DONOR QUANTITATIVE

```{r}
#works
# Input variance components and their SEs
Va    <- 0.02216302      # Genetic variance
Va_se <- 0.005141393     # SE of Va
Ve    <- 0.01752539      # Residual variance
Ve_se <- 0.001265809     # SE of Ve

# 1. Heritability estimate
h2 <- Va / (Va + Ve)

# 2. Delta method for SE of h2
denom <- (Va + Ve)^2

SE_h2 <- sqrt(
  (Ve / denom)^2 * Va_se^2 +
  (Va / denom)^2 * Ve_se^2
)

# 3. Z-ratio
z_ratio <- h2 / SE_h2

# 4. Print results
cat("Observed-scale heritability (h2):", round(h2, 4), "\n")
cat("Standard error of h2:            ", round(SE_h2, 4), "\n")
cat("Z-ratio:                         ", round(z_ratio, 2), "\n")

```

#host

```{r}

#ISSUES
# HOST QUANTITATIVE — NO DISTINCT

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- read_csv("hostquant.csv") %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# Convert columns to appropriate types
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# Prepare pedigree (no distinct used)
ped <- host_dat %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# First fit and update
mod_host <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)
mod_host <- update(mod_host)  # First update
mod_host <- update(mod_host)  # Second update ensures convergence + std.errors

# Now extract variance components with standard errors
vc <- summary(mod_host)$varcomp
print(vc)

# Extract components for heritability estimation
Va    <- vc["vm(offspring_id_host, ainv)", "component"]
SE_Va <- vc["vm(offspring_id_host, ainv)", "std.error"]
Ve    <- vc["units!R", "component"]


# Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)
mod_host <- update(mod_host)

# Extract variance components manually
vc <- summary(mod_host)$varcomp
print(vc)
Va <- vc["vm(offspring_id_host, ainv)", "component"]
Ve <- vc["units!R", "component"]

# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>% filter(host %in% unique(host_dat_clean$offspring_id_host))
blup_host$EBV <- blup_host$solution

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```

```{r}
# HOST QUANTITATIVE — CLEAN VERSION (NO DISTINCT, CONVERGED TWICE, NO HERITABILITY SECTION)

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- read_csv("hostquant.csv") %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# Convert columns to appropriate types
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# Prepare pedigree (no distinct used)
ped <- host_dat %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed = pred ~ 1,
  random = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data = host_dat_clean
)

# Update twice to ensure convergence and standard errors
mod_host <- update(mod_host)
mod_host <- update(mod_host)

# Extract and print variance components
vc <- summary(mod_host)$varcomp
print(vc)

# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>%
  filter(host %in% unique(host_dat_clean$offspring_id_host)) %>%
  mutate(EBV = solution)

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```

```{r}

# HOST QUANTITATIVE — CLEAN VERSION (NO DISTINCT, CONVERGED TWICE, NO HERITABILITY SECTION)

# Load required libraries
library(readr)
library(asreml)
library(nadiv)
library(dplyr)

# Load and clean data
host_dat <- read_csv("hostquant.csv") %>%
  filter(!is.na(offspring_id_host) & !is.na(dam_host) & !is.na(sire_host) & !is.na(pred))

# Convert columns to appropriate types
host_dat$offspring_id_host <- as.factor(host_dat$offspring_id_host)
host_dat$dam_host <- as.factor(host_dat$dam_host)
host_dat$sire_host <- as.factor(host_dat$sire_host)

# Prepare pedigree (no distinct used)
ped <- host_dat %>%
  select(ID = offspring_id_host, SIRE = sire_host, DAM = dam_host)
ped <- data.frame(lapply(ped, as.factor))
ainv <- ainverse(ped)

# Filter complete data for modeling
host_dat_clean <- host_dat %>%
  filter(complete.cases(pred, offspring_id_host))

# Fit ASReml Gaussian animal model
mod_host <- asreml(
  fixed    = pred ~ 1,
  random   = ~ vm(offspring_id_host, ainv),
  residual = ~ idv(units),
  data     = host_dat_clean
)


# Update twice to ensure convergence and standard errors
mod_host <- update(mod_host)
# mod_host <- update(mod_host)

# Extract and print variance components
vc <- summary(mod_host)$varcomp
print(vc)

# Extract BLUPs
blup_host <- as.data.frame(summary(mod_host, coef = TRUE)$coef.random)
blup_host$host <- gsub("^.*\\)_", "", rownames(blup_host))
blup_host <- blup_host %>%
  filter(host %in% unique(host_dat_clean$offspring_id_host)) %>%
  mutate(EBV = solution)

# View top 10 BLUPs
head(blup_host[order(-blup_host$EBV), ], 10)

```

```{r}
# From your model output:
Va <- 7.674054e-08   # Genetic variance component (very small)
Ve <- 0.04326674     # Residual variance

# Total phenotypic variance
Vp <- Va + Ve

# Heritability estimate
h2 <- Va / Vp
cat("Heritability estimate (h2):", h2, "\n")

#This is effectively zero, meaning:

#The host ID (additive genetic effect) explains almost none of the variance in the pred trait.

#The trait is likely dominated by environmental/residual variation in this dataset.


```

#Use a conservative assumed coefficient of variation (CV) #Assume a CV of 50%, i.e., standard error is 50% of the estimate.

```{r}
Va <- 7.674054e-08
Ve <- 0.04326674
SE_Va <- Va * 0.5   # 50% approximation
SE_Ve <- 0.003033001

Vp <- Va + Ve
denom <- (Vp)^2

# Delta method for SE of heritability
h2 <- Va / Vp
SE_h2 <- sqrt(
  (Ve / denom)^2 * SE_Va^2 +
  (Va / denom)^2 * SE_Ve^2
)
z_ratio <- h2 / SE_h2

cat("Heritability (h2):", h2, "\n")
cat("Approx. SE(h2):   ", SE_h2, "\n")
cat("Z-ratio:          ", z_ratio, "\n")



```

# graph

```{r}

# Load required libraries
library(ggplot2)
library(readr)

# Load the dataset
host_dat <- read_csv("hostquant.csv")

# Plot the distribution of 'pred'
ggplot(host_dat, aes(x = pred)) +
  geom_histogram(binwidth = 0.05, color = "black", fill = "steelblue", boundary = 0) +
  labs(
    title = "Distribution of Pearl Colour (pred values)",
    x = "Predicted Colour Score (pred)",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)



```

Assumption: Gold = pred \>= 0.8

Non-Gold = pred \< 0.8

```{r}
# Load required libraries
library(ggplot2)
library(readr)
library(dplyr)

# Load the dataset
host_dat <- read_csv("hostquant.csv")

# Define colour class
host_dat <- host_dat %>%
  mutate(colour_class = ifelse(pred >= 0.8, "Gold", "Non-Gold"))

# Plot overlaid histograms
ggplot(host_dat, aes(x = pred, fill = colour_class)) +
  geom_histogram(data = subset(host_dat, colour_class == "Non-Gold"),
                 binwidth = 0.05, alpha = 0.6, color = "black") +
  geom_histogram(data = subset(host_dat, colour_class == "Gold"),
                 binwidth = 0.05, alpha = 0.8, color = "black") +
  scale_fill_manual(values = c("Gold" = "goldenrod", "Non-Gold" = "grey60")) +
  labs(
    title = "Overlay of Gold vs Non-Gold Pearl Colour Scores",
    x = "Predicted Colour Score (pred)",
    y = "Count",
    fill = "Colour Class"
  ) +
  theme_minimal(base_size = 14)




```
